\chapter[Symptoms of Melancholy]{Symptoms, or Signs of Melancholy in the Body}\label{ch:symptoms}
{
%SECT. III. MEMB. I.
\lettrine{P}{arrhasius}, a painter of Athens, amongst those Olynthian captives
Philip of Macedon brought home to sell, \authorfootnote{2452}bought one very old man;
and when he had him at Athens, put him to extreme torture and torment,
the better by his example to express the pains and passions of his
Prometheus, whom he was then about to paint. I need not be so
barbarous, inhuman, curious, or cruel, for this purpose to torture any
poor melancholy man, their symptoms are plain, obvious and familiar,
there needs no such accurate observation or far-fetched object, they
delineate themselves, they voluntarily betray themselves, they are too
frequent in all places, I meet them still as I go, they cannot conceal
it, their grievances are too well known, I need not seek far to
describe them.

Symptoms therefore are either \authorfootnote{2453}universal or particular,
saith Gordonius, \bookcite{\textlatin{lib. med. cap. 19, part. 2}}, to persons,
to species; "some signs are secret, some manifest, some in the body, some in
the mind, and diversely vary, according to the inward or outward causes,"
Capivaccius: or from stars, according to Jovianus Pontanus,
\bookcite{\textlatin{de reb. caelest. lib. 10, cap. 13}}, and celestial
influences, or from the humours diversely mixed, Ficinus,
\bookcite{\textlatin{lib. 1, cap. 4, de sanit. tuenda}}: as they are hot, cold,
natural, unnatural, intended, or remitted, so will Aetius have \li{melancholica
deliria multiformia}, diversity of melancholy signs. Laurentius ascribes them
to their several temperatures, delights, natures, inclinations, continuance of
time, as they are simple or mixed with other diseases, as the causes are
divers, so must the signs be, almost infinite, Altomarus
\bookcite{\textlatin{cap. 7, art. med.}} And as wine produceth divers effects,
or that herb Tortocolla in \authorfootnote{2454}Laurentius, "which makes some
laugh, some weep, some sleep, some dance, some sing, some howl, some drink,
\etc{}" so doth this our melancholy humour work several signs in several
parties.

But to confine them, these general symptoms may be reduced to those of the body
or the mind. Those usual signs appearing in the bodies of such as are
melancholy, be these cold and dry, or they are hot and dry, as the humour is
more or less adust. From \authorfootnote{2455}these first qualities arise many
other second, as that of \authorfootnote{2456}colour, black, swarthy, pale,
ruddy, \etc{}, some are \li{impense rubri}, as Montaltus
\bookcite{\textlatin{cap. 16}} observes out of Galen, \bookcite{\textlatin{lib.
3, de locis affectis}}, very red and high coloured. Hippocrates in his book
\authorfootnote{2457}\bookcite{\textlatin{de insania et melan.}} reckons up
these signs, that they are \authorfootnote{2458}"lean, withered, hollow-eyed,
look old, wrinkled, harsh, much troubled with wind, and a griping in their
bellies, or bellyache, belch often, dry bellies and hard, dejected looks,
flaggy beards, singing of the ears, vertigo, light-headed, little or no sleep,
and that interrupt, terrible and fearful dreams," \authorfootnote{2459}\li{Anna
soror, quae, me suspensam insomnia terrent}? The same symptoms are repeated by
Melanelius in his book of melancholy collected out of Galen, Ruffus, Aetius, by
Rhasis, Gordonius, and all the juniors, \authorfootnote{2460}"continual, sharp,
and stinking belchings, as if their meat in their stomachs were putrefied, or
that they had eaten fish, dry bellies, absurd and interrupt dreams, and many
fantastical visions about their eyes, vertiginous, apt to tremble, and prone to
venery." \authorfootnote{2461}Some add palpitation of the heart, cold sweat, as
usual symptoms, and a leaping in many parts of the body, \li{saltum in multis
corporis partibus}, a kind of itching, saith Laurentius, on the superficies of
the skin, like a flea-biting sometimes. \authorfootnote{2462}Montaltus
\bookcite{\textlatin{cap. 21.}} puts fixed eyes and much twinkling of their
eyes for a sign, and so doth Avicenna, \li{oculos habentes palpitantes, trauli,
vehementer rubicundi}, \etc{}, \bookcite{\textlatin{lib. 3. Fen. 1. Tract. 4.
cap. 18.}} They stut most part, which he took out of Hippocrates' aphorisms.
\authorfootnote{2463}Rhasis makes "headache and a binding heaviness for a
principal token, much leaping of wind about the skin, as well as stutting, or
tripping in speech, \etc{}, hollow eyes, gross veins, and broad lips." To some
too, if they be far gone, mimical gestures are too familiar, laughing,
grinning, fleering, murmuring, talking to themselves, with strange mouths and
faces, inarticulate voices, exclamations, \etc{} And although they be commonly
lean, hirsute, uncheerful in countenance, withered, and not so pleasant to
behold, by reason of those continual fears, griefs, and vexations, dull, heavy,
lazy, restless, unapt to go about any business; yet their memories are most
part good, they have happy wits, and excellent apprehensions. Their hot and dry
brains make them they cannot sleep, \li{Ingentes habent et crebras vigilias}
(Arteus) mighty and often watchings, sometimes waking for a month, a year
together. \authorfootnote{2464}Hercules de Saxonia faithfully averreth, that he
hath heard his mother swear, she slept not for seven months together:
Trincavelius, \bookcite{\textlatin{Tom. 2. cons. 16.}} speaks of one that waked
50 days, and Skenkius hath examples of two years, and all without offence. In
natural actions their appetite is greater than their concoction, \li{multa
appetunt pauca digerunt} as Rhasis hath it, they covet to eat, but cannot
digest. And although they \authorfootnote{2465}"do eat much, yet they are lean,
ill-liking," saith Areteus, "withered and hard, much troubled with
costiveness," crudities, oppilations, spitting, belching, \etc{} Their pulse is
rare and slow, except it be of the \authorfootnote{2466}Carotides, which is
very strong; but that varies according to their intended passions or
perturbations, as Struthius hath proved at large,
\bookcite{\textlatin{Spigmaticae. artis l. 4. c. 13.}} To say truth, in such
chronic diseases the pulse is not much to be respected, there being so much
superstition in it, as \authorfootnote{2467}Crato notes, and so many
differences in Galen, that he dares say they may not be observed, or understood
of any man.

Their urine is most part pale, and low coloured, \li{urina pauca acris,
biliosa} (Areteus), not much in quantity; but this, in my judgment, is all out
as uncertain as the other, varying so often according to several persons,
habits, and other occasions not to be respected in chronic diseases.
\authorfootnote{2468}"Their melancholy excrements in some very much, in others
little, as the spleen plays his part," and thence proceeds wind, palpitation of
the heart, short breath, plenty of humidity in the stomach, heaviness of heart
and heartache, and intolerable stupidity and dullness of spirits. Their
excrements or stool hard, black to some and little. If the heart, brain, liver,
spleen, be misaffected, as usually they are, many inconveniences proceed from
them, many diseases accompany, as incubus, \authorfootnote{2469}apoplexy,
epilepsy, vertigo, those frequent wakings and terrible dreams,
\authorfootnote{2470}intempestive laughing, weeping, sighing, sobbing,
bashfulness, blushing, trembling, sweating, swooning, \etc{}
\authorfootnote{2471}All their senses are troubled, they think they see, hear,
smell, and touch that which they do not, as shall be proved in the following
discourse.

%\section{ SUBSECT. II.-- \emph{Symptoms or Signs in the Mind}.}
\section{Symptoms or Signs in the Mind.}

\subsection{Fear.}
\lettrine{A}{rculanus} \bookcite{\textlatin{in 9. Rhasis ad Almansor. cap. 16.}} will have
these symptoms to be infinite, as indeed they are, varying according to the
parties, "for scarce is there one of a thousand that dotes alike,"
\authorfootnote{2472}Laurentius \bookcite{\textlatin{c. 16.}} Some few of
greater note I will point at; and amongst the rest, fear and sorrow, which as
they are frequent causes, so if they persevere long, according to Hippocrates
\authorfootnote{2473}and Galen's aphorisms, they are most assured signs,
inseparable companions, and characters of melancholy; of present melancholy and
habituated, saith Montaltus \bookcite{\textlatin{cap. 11.}} and common to them
all, as the said Hippocrates, Galen, Avicenna, and all Neoterics hold. But as
hounds many times run away with a false cry, never perceiving themselves to be
at a fault, so do they. For Diocles of old, (whom Galen confutes,) and amongst
the juniors, \authorfootnote{2474}Hercules de Saxonia, with Lod. Mercatus
\bookcite{\textlatin{cap. 17. l. 1. de melan.}}, takes just exceptions, at this
aphorism of Hippocrates, 'tis not always true, or so generally to be
understood, "fear and sorrow are no common symptoms to all melancholy; upon
more serious consideration, I find some" (saith he) "that are not so at all.
Some indeed are sad, and not fearful; some fearful and not sad; some neither
fearful nor sad; some both." Four kinds he excepts, fanatical persons, such as
were Cassandra, Nanto, Nicostrata, Mopsus, Proteus, the sibyls, whom
\authorfootnote{2475}Aristotle confesseth to have been deeply melancholy.
Baptista Porta seconds him, \bookcite{\textlatin{Physiog. lib. 1, cap. 8}},
they were \li{atra bile perciti}: demoniacal persons, and such as speak strange
languages, are of this rank: some poets, such as laugh always, and think
themselves kings, cardinals, \etc{}, sanguine they are, pleasantly disposed
most part, and so continue. \authorfootnote{2476}Baptista Portia confines fear
and sorrow to them that are cold; but lovers, Sibyls, enthusiasts, he wholly
excludes. So that I think I may truly conclude, they are not always sad and
fearful, but usually so: and that \authorfootnote{2477}without a cause,
\li{timent de non timendis}, (Gordonius,) \li{quaeque momenti non sunt},
"although not all alike" (saith Altomarus), \authorfootnote{2478}"yet all
likely fear," \authorfootnote{2479}"some with an extraordinary and a mighty
fear," Areteus. \authorfootnote{2480}"Many fear death, and yet in a contrary
humour, make away themselves," Galen, \bookcite{\textlatin{lib. 3. de loc.
affec. cap. 7.}} Some are afraid that heaven will fall on their heads: some
they are damned, or shall be. \authorfootnote{2481}"They are troubled with
scruples of consciences, distrusting God's mercies, think they shall go
certainly to hell, the devil will have them, and make great lamentation," Jason
Pratensis. Fear of devils, death, that they shall be so sick, of some such or
such disease, ready to tremble at every object, they shall die themselves
forthwith, or that some of their dear friends or near allies are certainly
dead; imminent danger, loss, disgrace still torment others, \etc{}; that they
are all glass, and therefore will suffer no man to come near them: that they
are all cork, as light as feathers; others as heavy as lead; some are afraid
their heads will fall off their shoulders, that they have frogs in their
bellies, \etc{} \authorfootnote{2482}Montanus \bookcite{\textlatin{consil.
23}}, speaks of one "that durst not walk alone from home, for fear he should
swoon or die." A second \authorfootnote{2483}"fears every man he meets will rob
him, quarrel with him, or kill him." A third dares not venture to walk alone,
for fear he should meet the devil, a thief, be sick; fears all old women as
witches, and every black dog or cat he sees he suspecteth to be a devil, every
person comes near him is maleficiated, every creature, all intend to hurt him,
seek his ruin; another dares not go over a bridge, come near a pool, rock,
steep hill, lie in a chamber where cross beams are, for fear he be tempted to
hang, drown, or precipitate himself. If he be in a silent auditory, as at a
sermon, he is afraid he shall speak aloud at unawares, something indecent,
unfit to be said. If he be locked in a close room, he is afraid of being
stifled for want of air, and still carries biscuit, aquavitae, or some strong
waters about him, for fear of deliquiums, or being sick; or if he be in a
throng, middle of a church, multitude, where he may not well get out, though he
sit at ease, he is so misaffected. He will freely promise, undertake any
business beforehand, but when it comes to be performed, he dare not adventure,
but fears an infinite number of dangers, disasters, \etc{} Some are
\authorfootnote{2484}"afraid to be burned, or that the
\authorfootnote{2485}ground will sink under them, or
\authorfootnote{2486}swallow them quick, or that the king will call them in
question for some fact they never did (Rhasis \bookcite{\textlatin{cont.}}) and
that they shall surely be executed." The terror of such a death troubles them,
and they fear as much and are equally tormented in mind,
\authorfootnote{2487}"as they that have committed a murder, and are pensive
without a cause, as if they were now presently to be put to death." Plater,
\bookcite{\textlatin{cap. 3. de mentis alienat.}} They are afraid of some loss,
danger, that they shall surely lose their lives, goods, and all they have, but
why they know not. Trincavelius, \bookcite{\textlatin{consil. 13. lib. 1.}} had
a patient that would needs make away himself, for fear of being hanged, and
could not be persuaded for three years together, but that he had killed a man.
Plater, \bookcite{\textlatin{observat. lib. 1.}} hath two other examples of
such as feared to be executed without a cause. If they come in a place where a
robbery, theft, or any such offence hath been done, they presently fear they
are suspected, and many times betray themselves without a cause. Lewis XI., the
French king, suspected every man a traitor that came about him, durst trust no
officer. \li{Alii formidolosi omnium, alii quorundam} (Fracatorius
\bookcite{\textlatin{lib. 2. de Intellect.}}) \authorfootnote{2488}"some fear
all alike, some certain men, and cannot endure their companies, are sick in
them, or if they be from home." Some suspect \authorfootnote{2489}treason
still, others "are afraid of their \authorfootnote{2490}dearest and nearest
friends." (\li{Melanelius e Galeno, Ruffo, Aetio},) and dare not be alone in
the dark for fear of hobgoblins and devils: he suspects everything he hears or
sees to be a devil, or enchanted, and imagineth a thousand chimeras and
visions, which to his thinking he certainly sees, bugbears, talks with black
men, ghosts, goblins, \etc{}, \authorfootnote{2491}\li{Omnes se terrent aurae,
sonus excitat omnis.} Another through bashfulness, suspicion, and timorousness
will not be seen abroad, \authorfootnote{2492}"loves darkness as life, and
cannot endure the light," or to sit in lightsome places, his hat still in his
eyes, he will neither see nor be seen by his goodwill, Hippocrates,
\bookcite{\textlatin{lib. de Insania et Melancholia}}. He dare not come in
company for fear he should be misused, disgraced, overshoot himself in gesture
or speeches, or be sick; he thinks every man observes him, aims at him, derides
him, owes him malice. Most part \authorfootnote{2493}"they are afraid they are
bewitched, possessed, or poisoned by their enemies, and sometimes they suspect
their nearest friends: he thinks something speaks or talks within him, and he
belcheth of the poison." Christophorus a Vega, \bookcite{\textlatin{lib. 2.
cap. 1.}} had a patient so troubled, that by no persuasion or physic he could
be reclaimed. Some are afraid that they shall have every fearful disease they
see others have, hear of, or read, and dare not therefore hear or read of any
such subject, no not of melancholy itself, lest by applying to themselves that
which they hear or read, they should aggravate and increase it. If they see one
possessed, bewitched, an epileptic paroxysm, a man shaking with the palsy, or
giddy-headed, reeling or standing in a dangerous place, \etc{}, for many days
after it runs in their minds, they are afraid they shall be so too, they are in
like danger, as Perkins \bookcite{\textlatin{c. 12. sc. 12.}} well observes in
his Cases of Conscience and many times by violence of imagination they produce
it. They cannot endure to see any terrible object, as a monster, a man
executed, a carcase, hear the devil named, or any tragical relation seen, but
they quake for fear, \li{Hecatas somniare sibi videntur} (Lucian) they dream of
hobgoblins, and may not get it out of their minds a long time after: they apply
(as I have said) all they hear, see, read, to themselves; as
\authorfootnote{2494}Felix Plater notes of some young physicians, that study to
cure diseases, catch them themselves, will be sick, and appropriate all
symptoms they find related of others, to their own persons. And therefore
(\li{quod iterum moneo, licet nauseam paret lectori, malo decem potius verba,
decies repetita licet abundare, quam unum desiderari}) I would advise him that
is actually melancholy not to read this tract of Symptoms, lest he disquiet or
make himself for a time worse, and more melancholy than he was before.
Generally of them all take this, \li{de inanibus semper conqueruntur et
timent}, saith Aretius; they complain of toys, and fear
\authorfootnote{2495}without a cause, and still think their melancholy to be
most grievous, none so bad as they are, though it be nothing in respect, yet
never any man sure was so troubled, or in this sort. As really tormented and
perplexed, in as great an agony for toys and trifles (such things as they will
after laugh at themselves) as if they were most material and essential matters
indeed, worthy to be feared, and will not be satisfied. Pacify them for one,
they are instantly troubled with some other fear; always afraid of something
which they foolishly imagine or conceive to themselves, which never
peradventure was, never can be, never likely will be; troubled in mind upon
every small occasion, unquiet, still complaining, grieving, vexing, suspecting,
grudging, discontent, and cannot be freed so long as melancholy continues. Or
if their minds be more quiet for the present, and they free from foreign fears,
outward accidents, yet their bodies are out of tune, they suspect some part or
other to be amiss, now their head aches, heart, stomach, spleen, \etc{} is
misaffected, they shall surely have this or that disease; still troubled in
body, mind, or both, and through wind, corrupt fantasy, some accidental
distemper, continually molested. Yet for all this, as
\authorfootnote{2496}Jacchinus notes, "in all other things they are wise,
staid, discreet, and do nothing unbeseeming their dignity, person, or place,
this foolish, ridiculous, and childish fear excepted;" which so much, so
continually tortures and crucifies their souls, like a barking dog that always
bawls, but seldom bites, this fear ever molesteth, and so long as melancholy
lasteth, cannot be avoided.

Sorrow is that other character, and inseparable companion, as individual as
Saint Cosmus and Damian, \li{fidus Achates}, as all writers witness, a common
symptom, a continual, and still without any evident cause,
\authorfootnote{2497}\li{moerent omnes, et si roges eos reddere causam, non
possunt}: grieving still, but why they cannot tell: \li{Agelasti, moesti,
cogitabundi}, they look as if they had newly come forth of Trophonius' den. And
though they laugh many times, and seem to be extraordinary merry (as they will
by fits), yet extreme lumpish again in an instant, dull and heavy, \li{semel et
simul}, merry and sad, but most part sad: \authorfootnote{2498}\li{Si qua
placent, abeunt; inimica tenacius haerent}: sorrow sticks by them still
continually, gnawing as the vulture did \authorfootnote{2499}Titius' bowels,
and they cannot avoid it. No sooner are their eyes open, but after terrible and
troublesome dreams their heavy hearts begin to sigh: they are still fretting,
chafing, sighing, grieving, complaining, finding faults, repining, grudging,
weeping, \li{Heautontimorumenoi}, vexing themselves,
\authorfootnote{2500}disquieted in mind, with restless, unquiet thoughts,
discontent, either for their own, other men's or public affairs, such as
concern them not; things past, present, or to come, the remembrance of some
disgrace, loss, injury, abuses, \etc{} troubles them now being idle afresh, as
if it were new done; they are afflicted otherwise for some danger, loss, want,
shame, misery, that will certainly come, as they suspect and mistrust. Lugubris
Ate frowns upon them, insomuch that Areteus well calls it \li{angorem animi}, a
vexation of the mind, a perpetual agony. They can hardly be pleased, or eased,
though in other men's opinion most happy, go, tarry, run, ride,
\authorfootnote{2501}-- \li{post equitem sedet atra cura}: they cannot avoid
this feral plague, let them come in what company they will,
\authorfootnote{2502}\li{haeret leteri lethalis arundo}, as to a deer that is
struck, whether he run, go, rest with the herd, or alone, this grief remains:
irresolution, inconstancy, vanity of mind, their fear, torture, care, jealousy,
suspicion, \etc{}, continues, and they cannot be relieved. So
\authorfootnote{2503}he complained in the poet,

\begin{latin}
\begin{verse}%
Domum revertor moestus, atque animo fere\\*
Perturbato, atque incerto prae aegritudine,\\*
Assido, accurrunt servi: succos detrahunt,\\*
Video alios festinare, lectos sternere,\\*
Coenam apparare, pro se quisque sedulo\\*
Faciebant, quo illam mihi lenirent miseriam.\\!
\end{verse}%
\end{latin}
\translationrule%
\begin{verse}%
He came home sorrowful, and troubled in his mind,
his servants did all they possibly could to please him;
one pulled off his socks, another made ready his bed,
a third his supper, all did their utmost endeavours
to ease his grief, and exhilarate his person,
he was profoundly melancholy, he had lost his son,
\li{illud angebat}, that was his Cordolium,
his pain, his agony which could not be removed.
\end{verse}%

\subsection{\textlatin{Taedium vitae}.}
Hence it proceeds many times, that they are weary of their lives, and feral
thoughts to offer violence to their own persons come into their minds,
\lit{taedium vitae}{weariness of life} is a common symptom, \li{tarda fluunt, ingrataque tempora},
they are soon tired with all things; they will now tarry, now be gone; now in
bed they will rise, now up, then go to bed, now pleased, then again displeased;
now they like, by and by dislike all, weary of all, \li{sequitur nunc vivendi,
nunc moriendi cupido}, saith Aurelianus, \bookcite{\textlatin{lib. 1. cap. 6}},
but most part \authorfootnote{2504}\li{vitam damnant}, discontent, disquieted,
perplexed upon every light, or no occasion, object: often tempted, I say, to
make away themselves: \authorfootnote{2505}\li{Vivere nolunt, mori nesciunt}:
they cannot die, they will not live: they complain, weep, lament, and think
they lead a most miserable life, never was any man so bad, or so before, every
poor man they see is most fortunate in respect of them, every beggar that comes
to the door is happier than they are, they could be contented to change lives
with them, especially if they be alone, idle, and parted from their ordinary
company, molested, displeased, or provoked: grief, fear, agony, discontent,
wearisomeness, laziness, suspicion, or some such passion forcibly seizeth on
them. Yet by and by when they come in company again, which they like, or be
pleased, \li{suam sententiam rursus damnant, et vitae solatia delectantur}, as
Octavius Horatianus observes, \bookcite{\textlatin{lib. 2. cap. 5}}, they
condemn their former mislike, and are well pleased to live. And so they
continue, till with some fresh discontent they be molested again, and then they
are weary of their lives, weary of all, they will die, and show rather a
necessity to live, than a desire. Claudius the emperor, as
\authorfootnote{2506}Sueton describes him, had a spice of this disease, for
when he was tormented with the pain of his stomach, he had a conceit to make
away himself. Julius Caesar Claudinus, \bookcite{\textlatin{consil. 84.}} had a
Polonian to his patient, so affected, that through \authorfootnote{2507}fear
and sorrow, with which he was still disquieted, hated his own life, wished for
death every moment, and to be freed of his misery. Mercurialis another, and
another that was often minded to despatch himself, and so continued for many
years.

\subsection{Suspicion, Jealousy.}
Suspicion, and jealousy, are general symptoms: they are commonly distrustful,
apt to mistake, and amplify, \li{facile irascibiles},
\authorfootnote{2508}testy, pettish, peevish, and ready to snarl upon every
\authorfootnote{2509}small occasion, \li{cum amicissimis}, and without a cause,
\li{datum vel non datum}, it will be \li{scandalum acceptum}. If they speak in
jest, he takes it in good earnest. If they be not saluted, invited, consulted
with, called to counsel, \etc{}, or that any respect, small compliment, or
ceremony be omitted, they think themselves neglected, and contemned; for a time
that tortures them. If two talk together, discourse, whisper, jest, or tell a
tale in general, he thinks presently they mean him, applies all to himself,
\li{de se putat omnia dici}. Or if they talk with him, he is ready to
misconstrue every word they speak, and interpret it to the worst; he cannot
endure any man to look steadily on him, speak to him almost, laugh, jest, or be
familiar, or hem, or point, cough, or spit, or make a noise sometimes, \etc{}
\authorfootnote{2510}He thinks they laugh or point at him, or do it in disgrace
of him, circumvent him, contemn him; every man looks at him, he is pale, red,
sweats for fear and anger, lest somebody should observe him. He works upon it,
and long after this false conceit of an abuse troubles him. Montanus
\bookcite{\textlatin{consil. 22.}} gives instance in a melancholy Jew, that was
\li{Iracundior Adria}, so waspish and suspicious, \li{tam facile iratus}, that
no man could tell how to carry himself in his company.

\subsection{Inconstancy.}
Inconstant they are in all their actions, vertiginous, restless, unapt to
resolve of any business, they will and will not, persuaded to and fro upon
every small occasion, or word spoken: and yet if once they be resolved,
obstinate, hard to be reconciled. If they abhor, dislike, or distaste, once
settled, though to the better by odds, by no counsel, or persuasion, to be
removed. Yet in most things wavering, irresolute, unable to deliberate, through
fear, \li{faciunt, et mox facti poenitent (Areteus) avari, et paulo post
prodigi}. Now prodigal, and then covetous, they do, and by-and-by repent them
of that which they have done, so that both ways they are troubled, whether they
do or do not, want or have, hit or miss, disquieted of all hands, soon weary,
and still seeking change, restless, I say, fickle, fugitive, they may not abide
to tarry in one place long.

\begin{latin}
\begin{verse}%
Romae rus optans, absentem rusticus urbem\\*
Tollit ad astra---
\end{verse}%
\end{latin}
\attrib{\getauthornote{2511}}

no company long, or to persevere in any action or business.

\begin{latin}
\begin{verse}%
Et similis regum pueris, pappare minutum\\*
Poscit, et iratus mammae lallare recusat,
\end{verse}%
\end{latin}
\attrib{\getauthornote{2512}}

eftsoons pleased, and anon displeased, as a man that's bitten with fleas, or
that cannot sleep turns to and fro in his bed, their restless minds are tossed
and vary, they have no patience to read out a book, to play out a game or two,
walk a mile, sit an hour, \etc{}, erected and dejected in an instant; animated
to undertake, and upon a word spoken again discouraged.

\subsection{Passionate.}

Extreme passionate, \li{Quicquid volunt valde volunt}; and what they desire,
they do most furiously seek; anxious ever, and very solicitous, distrustful,
and timorous, envious, malicious, profuse one while, sparing another, but most
part covetous, muttering, repining, discontent, and still complaining,
grudging, peevish, \li{injuriarum tenaces}, prone to revenge, soon troubled,
and most violent in all their imaginations, not affable in speech, or apt to
vulgar compliment, but surly, dull, sad, austere;
\lit{cogitabundi}{thoughtfully} still, very intent, and as
\idxname{durer}[Albertus Durer] paints melancholy, like a sad woman leaning on
her arm with fixed looks, neglected habit,\authorfootnote{2513} \etc{}, held
therefore by some proud, soft, sottish, or half-mad, as the Abderites esteemed
of Democritus: and yet of a deep reach, excellent apprehension, judicious,
wise, and witty: for I am of that \authorfootnote{2514}nobleman's mind,
"Melancholy advanceth men's conceits, more than any humour whatsoever,"
improves their meditations more than any strong drink or sack. They are of
profound judgment in some things, although in others \li{non recte judicant
inquieti}, saith Fracastorius, \bookcite{\textlatin{lib. 2. de Intell}}. And as
Arculanus, \bookcite{\textlatin{c. 16. in 9. Rhasis}}, terms it, \li{Judicium
plerumque perversum, corrupti, cum judicant honesta inhonesta, et amicitiam
habent pro inimicitia}: they count honesty dishonesty, friends as enemies, they
will abuse their best friends, and dare not offend their enemies. Cowards most
part \li{et ad inferendam injuriam timidissimi}, saith Cardan,
\bookcite{\textlatin{lib. 8. cap. 4. de rerum varietate}}: loath to offend, and
if they chance to overshoot themselves in word or deed: or any small business
or circumstance be omitted, forgotten, they are miserably tormented, and frame
a thousand dangers and inconveniences to themselves, \li{ex musca elephantem},
if once they conceit it: overjoyed with every good rumour, tale, or prosperous
event, transported beyond themselves: with every small cross again, bad news,
misconceived injury, loss, danger, afflicted beyond measure, in great agony,
perplexed, dejected, astonished, impatient, utterly undone: fearful, suspicious
of all. Yet again, many of them desperate harebrains, rash, careless, fit to be
assassinates, as being void of all fear and sorrow, according to
\authorfootnote{2515}Hercules de Saxonia, "Most audacious, and such as dare
walk alone in the night, through deserts and dangerous places, fearing none."

\subsection{Amorous.}

"They are prone to love," and \authorfootnote{2516}easy to be taken;
\li{Propensi ad amorem et excandescentiam} (Montaltus \bookcite{\textlatin{cap.
21.}}) quickly enamoured, and dote upon all, love one dearly, till they see
another, and then dote on her, \li{Et hanc, et hanc, et illam, et omnes}, the
present moves most, and the last commonly they love best. Yet some again
\li{Anterotes}, cannot endure the sight of a woman, abhor the sex, as that same
melancholy \authorfootnote{2517}duke of Muscovy, that was instantly sick, if he
came but in sight of them; and that \authorfootnote{2518}Anchorite, that fell
into a cold palsy, when a woman was brought before him.

\subsection{Humorous.}

Humorous they are beyond all measure, sometimes profusely laughing,
extraordinarily merry, and then again weeping without a cause, (which is
familiar with many gentlewomen,) groaning, sighing, pensive, sad, almost
distracted, \li{multa absurda fingunt, et a ratione aliena} (saith
\authorfootnote{2519}Frambesarius), they feign many absurdities, vain, void of
reason: one supposeth himself to be a dog, cock, bear, horse, glass, butter,
\etc{} He is a giant, a dwarf, as strong as an hundred men, a lord, duke,
prince, \etc{} And if he be told he hath a stinking breath, a great nose, that
he is sick, or inclined to such or such a disease, he believes it eftsoons, and
peradventure by force of imagination will work it out. Many of them are
immovable, and fixed in their conceits, others vary upon every object, heard or
seen. If they see a stage-play, they run upon that a week after; if they hear
music, or see dancing, they have nought but bagpipes in their brain: if they
see a combat, they are all for arms. \authorfootnote{2520}If abused, an abuse
troubles them long after; if crossed, that cross, \etc{} Restless in their
thoughts and actions, continually meditating, \li{Velut aegri somnia, vanae
finguntur species}; more like dreams, than men awake, they fain a company of
antic, fantastical conceits, they have most frivolous thoughts, impossible to
be effected; and sometimes think verily they hear and see present before their
eyes such phantasms or goblins, they fear, suspect, or conceive, they still
talk with, and follow them. In fine, \li{cogitationes somniantibus similes, id
vigilant, quod alii somniant cogitabundi}, still, saith Avicenna, they wake, as
others dream, and such for the most part are their imaginations and conceits,
\authorfootnote{2521}absurd, vain, foolish toys, yet they are
\authorfootnote{2522}most curious and solicitous, continual, \li{et supra
modum}, Rhasis \bookcite{\textlatin{cont. lib. 1. cap. 9.}} \li{praemeditantur
de aliqua re}. As serious in a toy, as if it were a most necessary business, of
great moment, importance, and still, still, still thinking of it: \li{saeviunt
in se}, macerating themselves. Though they do talk with you, and seem to be
otherwise employed, and to your thinking very intent and busy, still that toy
runs in their mind, that fear, that suspicion, that abuse, that jealousy, that
agony, that vexation, that cross, that castle in the air, that crotchet, that
whimsy, that fiction, that pleasant waking dream, whatsoever it is. \li{Nec
interrogant} (saith \authorfootnote{2523}Fracastorius) \li{nec interrogatis
recte respondent}. They do not much heed what you say, their mind is on another
matter; ask what you will, they do not attend, or much intend that business
they are about, but forget themselves what they are saying, doing, or should
otherwise say or do, whither they are going, distracted with their own
melancholy thoughts. One laughs upon a sudden, another smiles to himself, a
third frowns, calls, his lips go still, he acts with his hand as he walks,
\etc{} 'Tis proper to all melancholy men, saith
\authorfootnote{2524}Mercurialis, \bookcite{\textlatin{con. 11.}} "What conceit
they have once entertained, to be most intent, violent, and continually about
it." \li{Invitas occurrit}, do what they may they cannot be rid of it, against
their wills they must think of it a thousand times over, \li{Perpetuo
molestantur nec oblivisci possunt}, they are continually troubled with it, in
company, out of company; at meat, at exercise, at all times and places,
\authorfootnote{2525}\li{non desinunt ea, quae, minime volunt, cogitare}, if it
be offensive especially, they cannot forget it, they may not rest or sleep for
it, but still tormenting themselves, \li{Sysiphi saxum volvunt sibi ipsis}, as
\authorfootnote{2526}Brunner observes, \li{Perpetua calamitas et miserabile
flagellum}.

\subsection{Bashfulness.}

\authorfootnote{2527}Crato, \authorfootnote{2528}Laurentius, and Fernelius, put
bashfulness for an ordinary symptom, \li{sabrusticus pudor}, or \li{vitiosus
pudor}, is a thing which much haunts and torments them. If they have been
misused, derided, disgraced, chidden, \etc{}, or by any perturbation of mind,
misaffected, it so far troubles them, that they become quite moped many times,
and so disheartened, dejected, they dare not come abroad, into strange
companies especially, or manage their ordinary affairs, so childish, timorous,
and bashful, they can look no man in the face; some are more disquieted in this
kind, some less, longer some, others shorter, by fits, \etc{}, though some on
the other side (according to \authorfootnote{2529}Fracastorius) be
\li{inverecundi et pertinaces}, impudent and peevish. But most part they are
very shamefaced, and that makes them with Pet. Blesensis, Christopher Urswick,
and many such, to refuse honours, offices, and preferments, which sometimes
fall into their mouths, they cannot speak, or put forth themselves as others
can, \li{timor hos, pudor impedit illos}, timorousness and bashfulness hinder
their proceedings, they are contented with their present estate, unwilling to
undertake any office, and therefore never likely to rise. For that cause they
seldom visit their friends, except some familiars: \li{pauciloqui}, of few
words, and oftentimes wholly silent. \authorfootnote{2530}Frambeserius, a
Frenchman, had two such patients, \li{omnino taciturnos}, their friends could
not get them to speak: Rodericus a Fonseca \bookcite{\textlatin{consult. tom.
2. 85. consil.}} gives instance in a young man, of twenty-seven years of age,
that was frequently silent, bashful, moped, solitary, that would not eat his
meat, or sleep, and yet again by fits apt to be angry, \etc{}

\subsection{Solitariness.}
Most part they are, as Plater notes, \li{desides, taciturni, aegre impulsi, nec
nisi coacti procedunt}, \etc{} they will scarce be compelled to do that which
concerns them, though it be for their good, so diffident, so dull, of small or
no compliment, unsociable, hard to be acquainted with, especially of strangers;
they had rather write their minds than speak, and above all things love
solitariness. \li{Ob voluptatem, an ob timorem soli sunt}? Are they so solitary
for pleasure (one asks,) or pain? for both; yet I rather think for fear and
sorrow, \etc{}

\begin{latin}
\begin{verse}%
Hinc metuunt cupiuntque, dolent fugiuntque, nec auras\\*
Respiciunt, clausi tenebris, et carcere caeco.
\end{verse}%
\end{latin}
\translationrule%
\begin{verse}%
Hence 'tis they grieve and fear, avoiding light,\\*
And shut themselves in prison dark from sight.
\end{verse}%
\attrib{\getauthornote{2531}}

As Bellerophon in Homer,\authorfootnote{2532}

\begin{latin}
\begin{verse}%
Qui miser in sylvis moerens errabat opacis,\\*
Ipse suum cor edens, hominum vestigia vitans.
\end{verse}%
\end{latin}
\translationrule%
\begin{verse}%
That wandered in the woods sad all alone,\\*
Forsaking men's society, making great moan.
\end{verse}%

They delight in floods and waters, desert places, to walk alone in orchards,
gardens, private walks, back lanes, averse from company, as Diogenes in his
tub, or Timon Misanthropus \authorfootnote{2533}, they abhor all companions at
last, even their nearest acquaintances and most familiar friends, for they have
a conceit (I say) every man observes them, will deride, laugh to scorn, or
misuse them, confining themselves therefore wholly to their private houses or
chambers, \li{fugiunt homines sine causa} (saith Rhasis) \li{et odio habent},
\bookcite{\textlatin{cont. l. 1. c. 9.}} they will diet themselves, feed and
live alone. It was one of the chiefest reasons why the citizens of Abdera
suspected Democritus to be melancholy and mad, because that, as Hippocrates
related in his Epistle to Philopaemenes, \authorfootnote{2534}"he forsook the
city, lived in groves and hollow trees, upon a green bank by a brook side, or
confluence of waters all day long, and all night." \li{Quae quidem} (saith he)
\li{plurimum atra bile vexatis et melancholicis eveniunt, deserta frequentant,
hominumque congressum aversantur}; \authorfootnote{2535}which is an ordinary
thing with melancholy men. The Egyptians therefore in their hieroglyphics
expressed a melancholy man by a hare sitting in her form, as being a most
timorous and solitary creature, Pierius \bookcite{\textlatin{Hieroglyph. l.
12.}} But this, and all precedent symptoms, are more or less apparent, as the
humour is intended or remitted, hardly perceived in some, or not all, most
manifest in others. Childish in some, terrible in others; to be derided in one,
pitied or admired in another; to him by fits, to a second continuate: and
howsoever these symptoms be common and incident to all persons, yet they are
the more remarkable, frequent, furious and violent in melancholy men. To speak
in a word, there is nothing so vain, absurd, ridiculous, extravagant,
impossible, incredible, so monstrous a chimera, so prodigious and strange,
\authorfootnote{2536}such as painters and poets durst not attempt, which they
will not really fear, feign, suspect and imagine unto themselves: and that
which \authorfootnote{2537}Lod. Vives said in a jest of a silly country fellow,
that killed his ass for drinking up the moon, \li{ut lunam mundo redderet}, you
may truly say of them in earnest; they will act, conceive all extremes,
contrarieties, and contradictions, and that in infinite varieties.
\li{Melancholici plane incredibilia sibi persuadent, ut vix omnibus saeculis
duo reperti sint, qui idem imaginati sint (Erastus de Lamiis)}, scarce two of
two thousand that concur in the same symptoms. The tower of Babel never yielded
such confusion of tongues, as the chaos of melancholy doth variety of symptoms.
There is in all melancholy \li{similitudo dissimilis}, like men's faces, a
disagreeing likeness still; and as in a river we swim in the same place, though
not in the same numerical water; as the same instrument affords several
lessons, so the same disease yields diversity of symptoms. Which howsoever they
be diverse, intricate, and hard to be confined, I will adventure yet in such a
vast confusion and generality to bring them into some order; and so descend to
particulars.

%SECT. III. MEMB. I. SUBSECT. III.-_Particular Symptoms from the influence of Stars, parts
\section[Symptoms from the influence of Stars]{Particular Symptoms from the influence of Stars, parts of the Body, and Humours.}

\lettrine{S}{ome} men have peculiar symptoms, according to their temperament
and crisis, which they had from the stars and those celestial influences,
variety of wits and dispositions, as Anthony Zara contends,
\bookcite{\textlatin{Anat. ingen. sect. 1. memb. 11, 12, 13, 14.}} \li{plurimum
irritant influentiae, caelestes, unde cientur animi aegritudines et morbi
corporum}. \authorfootnote{2538}One saith, diverse diseases of the body and
mind proceed from their influences, \authorfootnote{2539}as I have already
proved out of Ptolemy, Pontanus, Lemnius, Cardan, and others as they are
principal significators of manners, diseases, mutually irradiated, or lords of
the geniture, \etc{} Ptolomeus in his centiloquy, Hermes, or whosoever else the
author of that tract, attributes all these symptoms, which are in melancholy
men, to celestial influences: which opinion Mercurialis \bookcite{\textlatin{de
affect, lib. cap. 10.}} rejects; but, as I say, \authorfootnote{2540}Jovianus
Pontanus and others stiffly defend. That some are solitary, dull, heavy,
churlish; some again blithe, buxom, light, and merry, they ascribe wholly to
the stars. As if Saturn be predominant in his nativity, and cause melancholy in
his temperature, then \authorfootnote{2541}he shall be very austere, sullen,
churlish, black of colour, profound in his cogitations, full of cares,
miseries, and discontents, sad and fearful, always silent, solitary, still
delighting in husbandry, in woods, orchards, gardens, rivers, ponds, pools,
dark walks and close: \li{Cogitationes sunt velle aedificare, velle arbores
plantare, agros colere}, \etc{} To catch birds, fishes, \etc{} still contriving
and musing of such matters. If Jupiter domineers, they are more ambitious,
still meditating of kingdoms, magistracies, offices, honours, or that they are
princes, potentates, and how they would carry themselves, \etc{} If Mars, they
are all for wars, brave combats, monomachies, testy, choleric, harebrain, rash,
furious, and violent in their actions. They will feign themselves victors,
commanders, are passionate and satirical in their speeches, great braggers,
ruddy of colour. And though they be poor in show, vile and base, yet like
Telephus and Peleus in the \authorfootnote{2542}poet, \li{Ampullas jactant et
sesquipedalia verba}, "forget their swelling and gigantic words," their mouths
are full of myriads, and tetrarchs at their tongues' end. If the sun, they will
be lords, emperors, in conceit at least, and monarchs, give offices, honours,
\etc{} If Venus, they are still courting of their mistresses, and most apt to
love, amorously given, they seem to hear music, plays, see fine pictures,
dancers, merriments, and the like. Ever in love, and dote on all they see.
Mercurialists are solitary, much in contemplation, subtle, poets, philosophers,
and musing most part about such matters. If the moon have a hand, they are all
for peregrinations, sea voyages, much affected with travels, to discourse,
read, meditate of such things; wandering in their thoughts, diverse, much
delighting in waters, to fish, fowl, \etc{}

But the most immediate symptoms proceed from the temperature itself, and the
organical parts, as head, liver, spleen, mesaraic veins, heart, womb, stomach,
\etc{}, and most especially from distemperature of spirits (which, as
\authorfootnote{2543}Hercules de Saxonia contends, are wholly immaterial), or
from the four humours in those seats, whether they be hot or cold, natural,
unnatural, innate or adventitious, intended or remitted, simple or mixed, their
diverse mixtures, and several adustions, combinations, which may be as
diversely varied, as those \authorfootnote{2544}four first qualities in
\authorfootnote{2545}Clavius, and produce as many several symptoms and
monstrous fictions as wine doth effect, which as Andreas Bachius observes,
\bookcite{\textlatin{lib. 3. de vino, cap. 20.}} are infinite. Of greater note
be these.

If it be natural melancholy, as Lod. Mercatus, \bookcite{\textlatin{lib. 1.
cap. 17. de melan.}} T. Bright. \bookcite{\textlatin{c. 16.}} hath largely
described, either of the spleen, or of the veins, faulty by excess of quantity,
or thickness of substance, it is a cold and dry humour, as Montanus affirms,
\bookcite{\textlatin{consil. 26}} the parties are sad, timorous and fearful.
Prosper Calenus, in his book \bookcite{\textlatin{de atra bile}}, will have
them to be more stupid than ordinary, cold, heavy, solitary, sluggish. \li{Si
multam atram bilem et frigidam habent}. Hercules de Saxonia,
\bookcite{\textlatin{c. 19. l. 7.}} \authorfootnote{2546}"holds these that are
naturally melancholy, to be of a leaden colour or black," and so doth
Guianerius, \bookcite{\textlatin{c. 3. tract. 15.}} and such as think
themselves dead many times, or that they see, talk with black men, dead men,
spirits and goblins frequently, if it be in excess. These symptoms vary
according to the mixture of those four humours adust, which is unnatural
melancholy. For as Trallianus hath written, \bookcite{\textlatin{cap. 16. l.
7.}} \authorfootnote{2547}"There is not one cause of this melancholy, nor one
humour which begets, but divers diversely intermixed, from whence proceeds this
variety of symptoms:" and those varying again as they are hot or cold.
\authorfootnote{2548}"Cold melancholy" (saith Benedic. Vittorius Faventinus
\bookcite{\textlatin{pract. mag.}}) "is a cause of dotage, and more mild
symptoms, if hot or more adust, of more violent passions, and furies."
Fracastorius, \bookcite{\textlatin{l. 2. de intellect.}} will have us to
consider well of it, \authorfootnote{2549}"with what kind of melancholy every
one is troubled, for it much avails to know it; one is enraged by fervent heat,
another is possessed by sad and cold; one is fearful, shamefaced; the other
impudent and bold;" as Ajax, \li{Arma rapit superosque furens inpraelia
poscit}: quite mad or tending to madness. \li{Nunc hos, nunc impetit illos.}
Bellerophon on the other side, \li{solis errat male sanus in agris}, wanders
alone in the woods; one despairs, weeps, and is weary of his life, another
laughs, \etc{} All which variety is produced from the several degrees of heat
and cold, which \authorfootnote{2550}Hercules de Saxonia will have wholly
proceed from the distemperature of spirits alone, animal especially, and those
immaterial, the next and immediate causes of melancholy, as they are hot, cold,
dry, moist, and from their agitation proceeds that diversity of symptoms, which
he reckons up, in the \authorfootnote{2551}thirteenth chap. of his Tract of
Melancholy, and that largely through every part. Others will have them come
from the diverse adustion of the four humours, which in this unnatural
melancholy, by corruption of blood, adust choler, or melancholy natural,
\authorfootnote{2552}"by excessive distemper of heat turned, in comparison of
the natural, into a sharp lye by force of adustion, cause, according to the
diversity of their matter, diverse and strange symptoms," which T. Bright
reckons up in his following chapter. So doth \authorfootnote{2553}Arculanus,
according to the four principal humours adust, and many others.

For example, if it proceed from phlegm, (which is seldom and not so frequently
as the rest) \authorfootnote{2554}it stirs up dull symptoms, and a kind of
stupidity, or impassionate hurt: they are sleepy, saith
\authorfootnote{2555}Savanarola, dull, slow, cold, blockish, ass-like,
\li{Asininam melancholiam}, \authorfootnote{2556}Melancthon calls it, "they are
much given to weeping, and delight in waters, ponds, pools, rivers, fishing,
fowling, \etc{}" (Arnoldus \bookcite{\textlatin{breviar. 1. cap. 18.}}) They
are \authorfootnote{2557}pale of colour, slothful, apt to sleep, heavy;
\authorfootnote{2558}much troubled with headache, continual meditation, and
muttering to themselves; they dream of waters, \authorfootnote{2559}that they
are in danger of drowning, and fear such things, Rhasis. They are fatter than
others that are melancholy, of a muddy complexion, apter to spit,
\authorfootnote{2560}sleep, more troubled with rheum than the rest, and have
their eyes still fixed on the ground. Such a patient had Hercules de Saxonia, a
widow in Venice, that was fat and very sleepy still; Christophorus a Vega
another affected in the same sort. If it be inveterate or violent, the symptoms
are more evident, they plainly denote and are ridiculous to others, in all
their gestures, actions, speeches; imagining impossibilities, as he in
Christophorus a Vega, that thought he was a tun of wine,
\authorfootnote{2561}and that Siennois, that resolved within himself not to
piss, for fear he should drown all the town.

If it proceed from blood adust, or that there be a mixture of blood in it,
\authorfootnote{2562}"such are commonly ruddy of complexion, and
high-coloured," according to Salust. Salvianus, and Hercules de Saxonia. And as
Savanarola, Vittorius Faventinus Emper. farther adds, \authorfootnote{2563}"the
veins of their eyes be red, as well as their faces." They are much inclined to
laughter, witty and merry, conceited in discourse, pleasant, if they be not far
gone, much given to music, dancing, and to be in women's company. They meditate
wholly on such things, and think \authorfootnote{2564}"they see or hear plays,
dancing, and suchlike sports" (free from all fear and sorrow, as
\authorfootnote{2565}Hercules de Saxonia supposeth.) If they be more strongly
possessed with this kind of melancholy, Arnoldus adds,
\bookcite{\textlatin{Breviar. lib. 1. cap. 18.}} Like him of Argos in the Poet,
that sate laughing \authorfootnote{2566}all day long, as if he had been at a
theatre. Such another is mentioned by \authorfootnote{2567}Aristotle, living at
Abydos, a town of Asia Minor, that would sit after the same fashion, as if he
had been upon a stage, and sometimes act himself; now clap his hands, and
laugh, as if he had been well pleased with the sight. Wolfius relates of a
country fellow called Brunsellius, subject to this humour,
\authorfootnote{2568}"that being by chance at a sermon, saw a woman fall off
from a form half asleep, at which object most of the company laughed, but he
for his part was so much moved, that for three whole days after he did nothing
but laugh, by which means he was much weakened, and worse a long time
following." Such a one was old Sophocles, and Democritus himself had \li{hilare
delirium}, much in this vein. Laurentius \bookcite{\textlatin{cap. 3. de
melan.}} thinks this kind of melancholy, which is a little adust with some
mixture of blood, to be that which Aristotle meant, when he said melancholy men
of all others are most witty, which causeth many times a divine ravishment, and
a kind of \li{enthusiasmus}, which stirreth them up to be excellent
philosophers, poets, prophets, \etc{} Mercurialis, \bookcite{\textlatin{consil.
110.}} gives instance in a young man his patient, sanguine melancholy,
\authorfootnote{2569}"of a great wit, and excellently learned."

If it arise from choler adust, they are bold and impudent, and of a more
harebrain disposition, apt to quarrel, and think of such things, battles,
combats, and their manhood, furious; impatient in discourse, stiff,
irrefragable and prodigious in their tenets; and if they be moved, most
violent, outrageous, \authorfootnote{2570}ready to disgrace, provoke any, to
kill themselves and others; Arnoldus adds, stark mad by fits,
\authorfootnote{2571}"they sleep little, their urine is subtle and fiery."
(Guianerius.) "In their fits you shall hear them speak all manner of languages,
Hebrew, Greek, and Latin, that never were taught or knew them before."
Apponensis in \bookcite{\textlatin{com. in Pro. sec. 30.}} speaks of a mad
woman that spake excellent good Latin: and Rhasis knew another, that could
prophecy in her fit, and foretell things truly to come.
\authorfootnote{2572}Guianerius had a patient could make Latin verses when the
moon was combust, otherwise illiterate. Avicenna and some of his adherents will
have these symptoms, when they happen, to proceed from the devil, and that they
are rather \li{demoniaci}, possessed, than mad or melancholy, or both together,
as Jason Pratensis thinks, \li{Immiscent se mali genii}, \etc{} but most
ascribe it to the humour, which opinion Montaltus \bookcite{\textlatin{cap.
21.}} stiffly maintains, confuting Avicenna and the rest, referring it wholly
to the quality and disposition of the humour and subject. Cardan
\bookcite{\textlatin{de rerum var. lib. 8. cap. 10.}} holds these men of all
others fit to be assassins, bold, hardy, fierce, and adventurous, to undertake
anything by reason of their choler adust. \authorfootnote{2573}"This humour,
saith he, prepares them to endure death itself, and all manner of torments with
invincible courage, and 'tis a wonder to see with what alacrity they will
undergo such tortures," \li{ut supra naturam res videatur}: he ascribes this
generosity, fury, or rather stupidity, to this adustion of choler and
melancholy: but I take these rather to be mad or desperate, than properly
melancholy; for commonly this humour so adust and hot, degenerates into
madness.

If it come from melancholy itself adust, those men, saith Avicenna,
\authorfootnote{2574}"are usually sad and solitary, and that continually, and
in excess, more than ordinarily suspicious more fearful, and have long, sore,
and most corrupt imaginations;" cold and black, bashful, and so solitary, that
as \authorfootnote{2575}Arnoldus writes, "they will endure no company, they
dream of graves still, and dead men, and think themselves bewitched or dead:"
if it be extreme, they think they hear hideous noises, see and talk
\authorfootnote{2576}"with black men, and converse familiarly with devils, and
such strange chimeras and visions," (Gordonius) or that they are possessed by
them, that somebody talks to them, or within them. \li{Tales melancholici
plerumque daemoniaci}, Montaltus \bookcite{\textlatin{consil. 26. ex
Avicenna}}. Valescus de Taranta had such a woman in cure,
\authorfootnote{2577}"that thought she had to do with the devil:" and Gentilis
Fulgosus \bookcite{\textlatin{quaest. 55.}} writes that he had a melancholy
friend, that \authorfootnote{2578}"had a black man in the likeness of a
soldier" still following him wheresoever he was. Laurentius
\bookcite{\textlatin{cap. 7.}} hath many stories of such as have thought
themselves bewitched by their enemies; and some that would eat no meat as being
dead. \authorfootnote{2579}\emph{Anno} 1550 an advocate of Paris fell into such
a melancholy fit, that he believed verily he was dead, he could not be
persuaded otherwise, or to eat or drink, till a kinsman of his, a scholar of
Bourges, did eat before him dressed like a corse. The story, saith Serres, was
acted in a comedy before Charles the Ninth. Some think they are beasts, wolves,
hogs, and cry like dogs, foxes, bray like asses, and low like kine, as King
Praetus' daughters. \authorfootnote{2580}Hildesheim
\bookcite{\textlatin{spicel. 2. de mania}}, hath an example of a Dutch baron so
affected, and Trincavelius \bookcite{\textlatin{lib. 1. consil. 11.}} another
of a nobleman in his country, \authorfootnote{2581}"that thought he was
certainly a beast, and would imitate most of their voices," with many such
symptoms, which may properly be reduced to this kind.

If it proceed from the several combinations of these four humours, or spirits,
Herc. de Saxon. adds hot, cold, dry, moist, dark, confused, settled,
constringed, as it participates of matter, or is without matter, the symptoms
are likewise mixed. One thinks himself a giant, another a dwarf. One is heavy
as lead, another is as light as a feather. Marcellus Donatus
\bookcite{\textlatin{l. 2. cap. 41.}} makes mention out of Seneca, of one
Seneccio, a rich man, \authorfootnote{2582}"that thought himself and everything
else he had, great: great wife, great horses, could not abide little things,
but would have great pots to drink in, great hose, and great shoes bigger than
his feet." Like her in \authorfootnote{2583}Trallianus, that supposed she
"could shake all the world with her finger," and was afraid to clinch her hand
together, lest she should crush the world like an apple in pieces: or him in
Galen, that thought he was \authorfootnote{2584}Atlas, and sustained heaven
with his shoulders. Another thinks himself so little, that he can creep into a
mouse-hole: one fears heaven will fall on his head: a second is a cock; and
such a one, \authorfootnote{2585}Guianerius saith he saw at Padua, that would
clap his hands together and crow. \authorfootnote{2586}Another thinks he is a
nightingale, and therefore sings all the night long; another he is all glass, a
pitcher, and will therefore let nobody come near him, and such a one
\authorfootnote{2587}Laurentius gives out upon his credit, that he knew in
France. Christophorus a Vega \bookcite{\textlatin{cap. 3. lib. 14.}} Skenkius
and Marcellus Donatus \bookcite{\textlatin{l. 2. cap. 1.}} have many such
examples, and one amongst the rest of a baker in Ferrara that thought he was
composed of butter, and durst not sit in the sun, or come near the fire for
fear of being melted: of another that thought he was a case of leather, stuffed
with wind. Some laugh, weep; some are mad, some dejected, moped, in much agony,
some by fits, others continuate, \etc{} Some have a corrupt ear, they think
they hear music, or some hideous noise as their phantasy conceives, corrupt
eyes, some smelling, some one sense, some another. \authorfootnote{2588}Lewis
the Eleventh had a conceit everything did stink about him, all the odoriferous
perfumes they could get, would not ease him, but still he smelled a filthy
stink. A melancholy French poet in \authorfootnote{2589}Laurentius, being sick
of a fever, and troubled with waking, by his physicians was appointed to use
\li{unguentum populeum} to anoint his temples; but he so distasted the smell of
it, that for many years after, all that came near him he imagined to scent of
it, and would let no man talk with him but aloof off, or wear any new clothes,
because he thought still they smelled of it; in all other things wise and
discreet, he would talk sensibly, save only in this. A gentleman in Limousin,
saith Anthony Verdeur, was persuaded he had but one leg, affrighted by a wild
boar, that by chance struck him on the leg; he could not be satisfied his leg
was sound (in all other things well) until two Franciscans by chance coming
that way, fully removed him from the conceit. \li{Sed abunde fabularum
audivimus},--enough of story-telling.

%SECT. III. MEMB. I. SUBSECT. IV.-_Symptoms from Education, Custom, continuance of Time, our Condition, mixed with other Diseases, by Fits, Inclination, \&c._
\section[Symptoms from Education and Custom]{Symptoms from Education, Custom, continuance of Time, our Condition, mixed with other Diseases, by Fits, Inclination, \&c.}

\lettrine{A}{nother} great occasion of the variety of these symptoms proceeds from custom,
discipline, education, and several inclinations, \authorfootnote{2590}"this
humour will imprint in melancholy men the objects most answerable to their
condition of life, and ordinary actions, and dispose men according to their
several studies and callings." If an ambitious man become melancholy, he
forthwith thinks he is a king, an emperor, a monarch, and walks alone, pleasing
himself with a vain hope of some future preferment, or present as he supposeth,
and withal acts a lord's part, takes upon him to be some statesman or
magnifico, makes conges, gives entertainment, looks big, \etc{} Francisco
Sansovino records of a melancholy man in Cremona, that would not be induced to
believe but that he was pope, gave pardons, made cardinals, \etc{}
\authorfootnote{2591}Christophorus a Vega makes mention of another of his
acquaintance, that thought he was a king, driven from his kingdom, and was very
anxious to recover his estate. A covetous person is still conversant about
purchasing of lands and tenements, plotting in his mind how to compass such and
such manors, as if he were already lord of, and able to go through with it; all
he sees is his, \li{re} or \li{spe}, he hath devoured it in hope, or else in
conceit esteems it his own: like him in \authorfootnote{2592}Athenaeus, that
thought all the ships in the haven to be his own. A lascivious \li{inamorato}
plots all the day long to please his mistress, acts and struts, and carries
himself as if she were in presence, still dreaming of her, as Pamphilus of his
Glycerium, or as some do in their morning sleep. \authorfootnote{2593}Marcellus
Donatus knew such a gentlewoman in Mantua, called Elionora Meliorina, that
constantly believed she was married to a king, and \authorfootnote{2594}"would
kneel down and talk with him, as if he had been there present with his
associates; and if she had found by chance a piece of glass in a muck-hill or
in the street, she would say that it was a jewel sent from her lord and
husband." If devout and religious, he is all for fasting, prayer, ceremonies,
alms, interpretations, visions, prophecies, revelations,
\authorfootnote{2595}he is inspired by the Holy Ghost, full of the spirit: one
while he is saved, another while damned, or still troubled in mind for his
sins, the devil will surely have him, \etc{} more of these in the third
partition of love-melancholy. \authorfootnote{2596}A scholar's mind is busied
about his studies, he applauds himself for that he hath done, or hopes to do,
one while fearing to be out in his next exercise, another while contemning all
censures; envies one, emulates another; or else with indefatigable pains and
meditation, consumes himself. So of the rest, all which vary according to the
more remiss and violent impression of the object, or as the humour itself is
intended or remitted. For some are so gently melancholy, that in all their
carriage, and to the outward apprehension of others it can hardly be discerned,
yet to them an intolerable burden, and not to be endured.
\authorfootnote{2597}\li{Quaedam occulta quaedam manifesta}, some signs are
manifest and obvious to all at all times, some to few, or seldom, or hardly
perceived; let them keep their own council, none will take notice or suspect
them. "They do not express in outward show their depraved imaginations," as
\authorfootnote{2598}Hercules de Saxonia observes, "but conceal them wholly to
themselves, and are very wise men, as I have often seen; some fear, some do not
fear at all, as such as think themselves kings or dead, some have more signs,
some fewer, some great, some less," some vex, fret, still fear, grieve, lament,
suspect, laugh, sing, weep, chafe, \etc{} by fits (as I have said) or more
during and permanent. Some dote in one thing, are most childish, and
ridiculous, and to be wondered at in that, and yet for all other matters most
discreet and wise. To some it is in disposition, to another in habit; and as
they write of heat and cold, we may say of this humour, one is
\li{melancholicus ad octo}, a second two degrees less, a third halfway. 'Tis
superparticular, \li{sesquialtera, sesquitertia}, and \li{superbipartiens
tertias, quintas Melancholiae}, \etc{} all those geometrical proportions are
too little to express it. \authorfootnote{2599}"It comes to many by fits, and
goes; to others it is continuate:" many (saith \authorfootnote{2600}Faventinus)
"in spring and fall only are molested," some once a year, as that Roman
\authorfootnote{2601}Galen speaks of: \authorfootnote{2602}one, at the
conjunction of the moon alone, or some unfortunate aspects, at such and such
set hours and times, like the sea-tides, to some women when they be with child,
as \authorfootnote{2603}Plater notes, never otherwise: to others 'tis settled
and fixed; to one led about and variable still by that \li{ignis fatuus} of
phantasy, like an \li{arthritis} or running gout, 'tis here and there, and in
every joint, always molesting some part or other; or if the body be free, in a
myriad of forms exercising the mind. A second once peradventure in his life
hath a most grievous fit, once in seven years, once in five years, even to the
extremity of madness, death, or dotage, and that upon, some feral accident or
perturbation, terrible object, and for a time, never perhaps so before, never
after. A third is moved upon all such troublesome objects, cross fortune,
disaster, and violent passions, otherwise free, once troubled in three or four
years. A fourth, if things be to his mind, or he in action, well pleased, in
good company, is most jocund, and of a good complexion: if idle, or alone, a la
mort, or carried away wholly with pleasant dreams and phantasies, but if once
crossed and displeased,

\begin{latin}
\begin{verse}%
Pectore concipiet nil nisi triste suo;
\end{verse}%
\end{latin}
\translationrule
\begin{verse}%
He will imagine naught save sadness in his heart;
\end{verse}%

his countenance is altered on a sudden, his heart heavy, irksome thoughts
crucify his soul, and in an instant he is moped or weary of his life, he will
kill himself. A fifth complains in his youth, a sixth in his middle age, the
last in his old age.

Generally thus much we may conclude of melancholy; that it is
\authorfootnote{2604}most pleasant at first, I say, \li{mentis gratissimus
error}, \authorfootnote{2605}a most delightsome humour, to be alone, dwell
alone, walk alone, meditate, lie in bed whole days, dreaming awake as it were,
and frame a thousand fantastical imaginations unto themselves. They are never
better pleased than when they are so doing, they are in paradise for the time,
and cannot well endure to be interrupt; with him in the poet,
\authorfootnote{2606}\li{pol me occidistis amici, non servastis ait}? you have
undone him, he complains, if you trouble him: tell him what inconvenience will
follow, what will be the event, all is one, \li{canis ad vomitum},
\authorfootnote{2607}'tis so pleasant he cannot refrain. He may thus continue
peradventure many years by reason of a strong temperature, or some mixture of
business, which may divert his cogitations: but at the last \li{laesa
imaginatio}, his phantasy is crazed, and now habituated to such toys, cannot
but work still like a fate, the scene alters upon a sudden, fear and sorrow
supplant those pleasing thoughts, suspicion, discontent, and perpetual anxiety
succeed in their places; so by little and little, by that shoeing-horn of
idleness, and voluntary solitariness, melancholy this feral fiend is drawn on,
\authorfootnote{2608}\li{et quantum vertice ad auras Aethereas, tantum radice
in Tartara tendit}, "extending up, by its branches, so far towards Heaven, as,
by its roots, it does down towards Tartarus;" it was not so delicious at first,
as now it is bitter and harsh; a cankered soul macerated with cares and
discontents, \li{taedium vitae}, impatience, agony, inconstancy, irresolution,
precipitate them unto unspeakable miseries. They cannot endure company, light,
or life itself, some unfit for action, and the like. \authorfootnote{2609}Their
bodies are lean and dried up, withered, ugly, their looks harsh, very dull, and
their souls tormented, as they are more or less entangled, as the humour hath
been intended, or according to the continuance of time they have been troubled.

To discern all which symptoms the better, \authorfootnote{2610}Rhasis the
Arabian makes three degrees of them. The first is, \li{falsa cogitatio}, false
conceits and idle thoughts: to misconstrue and amplify, aggravating everything
they conceive or fear; the second is, \li{falso cogitata loqui}, to talk to
themselves, or to use inarticulate incondite voices, speeches, obsolete
gestures, and plainly to utter their minds and conceits of their hearts, by
their words and actions, as to laugh, weep, to be silent, not to sleep, eat
their meat, \etc{}: the third is to put in practice \authorfootnote{2611}that
which they think or speak. Savanarola, \bookcite{\textlatin{Rub. 11. tract. 8.
cap. 1. de aegritudine}}, confirms as much, \authorfootnote{2612}"when he
begins to express that in words, which he conceives in his heart, or talks
idly, or goes from one thing to another," which \authorfootnote{2613}Gordonius
calls \li{nec caput habentia, nec caudam}, ("having neither head nor tail,") he
is in the middle way: \authorfootnote{2614}"but when he begins to act it
likewise, and to put his fopperies in execution, he is then in the extent of
melancholy, or madness itself." This progress of melancholy you shall easily
observe in them that have been so affected, they go smiling to themselves at
first, at length they laugh out; at first solitary, at last they can endure no
company: or if they do, they are now dizzards, past sense and shame, quite
moped, they care not what they say or do, all their actions, words, gestures,
are furious or ridiculous. At first his mind is troubled, he doth not attend
what is said, if you tell him a tale, he cries at last, what said you? but in
the end he mutters to himself, as old women do many times, or old men when they
sit alone, upon a sudden they laugh, whoop, halloo, or run away, and swear they
see or hear players, \authorfootnote{2615}devils, hobgoblins, ghosts, strike,
or strut, \etc{}, grow humorous in the end; like him in the poet, \li{saepe
ducentos, saepe decem servos}, ("at one time followed by two hundred servants,
at another only by ten") he will dress himself, and undress, careless at last,
grows insensible, stupid, or mad. \authorfootnote{2616}He howls like a wolf,
barks like a dog, and raves like Ajax and Orestes, hears music and outcries,
which no man else hears. As \authorfootnote{2617}he did whom Amatus Lusitanus
mentioneth \bookcite{\textlatin{cent. 3, cura. 55}}, or that woman in
\authorfootnote{2618}Springer, that spake many languages, and said she was
possessed: that farmer in \authorfootnote{2619}Prosper Calenius, that disputed
and discoursed learnedly in philosophy and astronomy, with Alexander Achilles
his master, at Bologna, in Italy. But of these I have already spoken.

Who can sufficiently speak of these symptoms, or prescribe rules to comprehend
them? as Echo to the painter in Ausonius, \li{vane quid affectas}, \etc{},
foolish fellow; what wilt? if you must needs paint me, paint a voice, \li{et
similem si vis pingere, pinge sonum}; if you will describe melancholy, describe
a fantastical conceit, a corrupt imagination, vain thoughts and different,
which who can do? The four and twenty letters make no more variety of words in
diverse languages, than melancholy conceits produce diversity of symptoms in
several persons. They are irregular, obscure, various, so infinite, Proteus
himself is not so diverse, you may as well make the moon a new coat, as a true
character of a melancholy man; as soon find the motion of a bird in the air, as
the heart of man, a melancholy man. They are so confused, I say, diverse,
intermixed with other diseases. As the species be confounded (which
\authorfootnote{2620}I have showed) so are the symptoms; sometimes with
headache, cachexia, dropsy, stone; as you may perceive by those several
examples and illustrations, collected by \authorfootnote{2621}Hildesheim
\bookcite{\textlatin{spicel. 2.}} Mercurialis \bookcite{\textlatin{consil. 118.
cap. 6 and 11.}} with headache, epilepsy, priapismus. Trincavelius
\bookcite{\textlatin{consil. 12. lib. 1. consil. 49.}} with gout: \li{caninus
appetitus}. Montanus \bookcite{\textlatin{consil. 26, \etc{} 23, 234, 249}},
with falling-sickness, headache, vertigo, lycanthropia, \etc{} J. Caesar
Claudinus \bookcite{\textlatin{consult. 4. consult. 89 and 116.}} with gout,
agues, haemorrhoids, stone, \etc{}, who can distinguish these melancholy
symptoms so intermixed with others, or apply them to their several kinds,
confine them into method? 'Tis hard I confess, yet I have disposed of them as I
could, and will descend to particularise them according to their species. For
hitherto I have expatiated in more general lists or terms, speaking
promiscuously of such ordinary signs, which occur amongst writers. Not that
they are all to be found in one man, for that were to paint a monster or
chimera, not a man: but some in one, some in another, and that successively or
at several times.

Which I have been the more curious to express and report; not to upbraid any
miserable man, or by way of derision, (I rather pity them,) but the better to
discern, to apply remedies unto them; and to show that the best and soundest of
us all is in great danger; how much we ought to fear our own fickle estates,
remember our miseries and vanities, examine and humiliate ourselves, seek to
God, and call to Him for mercy, that needs not look for any rods to scourge
ourselves, since we carry them in our bowels, and that our souls are in a
miserable captivity, if the light of grace and heavenly truth doth not shine
continually upon us: and by our discretion to moderate ourselves, to be more
circumspect and wary in the midst of these dangers.

%SECT. III. MEMB. II. SUBSECT. I.-_Symptoms of Head-Melancholy_.
\section{Symptoms of Head-Melancholy.}

\lettrine[ante="]{I}{f} \authorfootnote{2622}no symptoms appear about the stomach, nor
the blood be misaffected, and fear and sorrow continue, it is to be thought the
brain itself is troubled, by reason of a melancholy juice bred in it, or
otherwise conveyed into it, and that evil juice is from the distemperature of
the part, or left after some inflammation," thus far Piso. But this is not
always true, for blood and hypochondries both are often affected even in
head-melancholy. \authorfootnote{2623}Hercules de Saxonia differs here from the
common current of writers, putting peculiar signs of head-melancholy, from the
sole distemperature of spirits in the brain, as they are hot, cold, dry, moist,
"all without matter from the motion alone, and tenebrosity of spirits;" of
melancholy which proceeds from humours by adustion, he treats apart, with their
several symptoms and cures. The common signs, if it be by essence in the head,
"are ruddiness of face, high sanguine complexion, most part \li{rubore
saturato}," \authorfootnote{2624}one calls it, a bluish, and sometimes full of
pimples, with red eyes. Avicenna \bookcite{\textlatin{l. 3, Fen. 2, Tract. 4,
c. 18.}} Duretus and others out of Galen, \bookcite{\textlatin{de affect. l. 3,
c. 6.}} \authorfootnote{2625}Hercules de Saxonia to this of redness of face,
adds "heaviness of the head, fixed and hollow eyes." \authorfootnote{2626}"If
it proceed from dryness of the brain, then their heads will be light,
vertiginous, and they most apt to wake, and to continue whole months together
without sleep. Few excrements in their eyes and nostrils, and often bald by
reason of excess of dryness," Montaltus adds, \bookcite{\textlatin{c. 17.}} If
it proceed from moisture: dullness, drowsiness, headache follows; and as
Salust. Salvianus, \bookcite{\textlatin{c. 1, l. 2}}, out of his own experience
found, epileptical, with a multitude of humours in the head. They are very
bashful, if ruddy, apt to blush, and to be red upon all occasions,
\li{praesertim si metus accesserit}. But the chiefest symptom to discern this
species, as I have said, is this, that there be no notable signs in the
stomach, hypochondries, or elsewhere, \li{digna}, as
\authorfootnote{2627}Montaltus terms them, or of greater note, because
oftentimes the passions of the stomach concur with them. Wind is common to all
three species, and is not excluded, only that of the hypochondries is
\authorfootnote{2628}more windy than the rest, saith Hollerius. Aetius
\bookcite{\textlatin{tetrab. l. 2, sc. 2, c. 9 and 10}}, maintains the same,
\authorfootnote{2629}if there be more signs, and more evident in the head than
elsewhere, the brain is primarily affected, and prescribes head-melancholy to
be cured by meats amongst the rest, void of wind, and good juice, not excluding
wind, or corrupt blood, even in head-melancholy itself: but these species are
often confounded, and so are their symptoms, as I have already proved. The
symptoms of the mind are superfluous and continual cogitations;
\authorfootnote{2630}"for when the head is heated, it scorcheth the blood, and
from thence proceed melancholy fumes, which trouble the mind," Avicenna. They
are very choleric, and soon hot, solitary, sad, often silent, watchful,
discontent, Montaltus, \bookcite{\textlatin{cap. 24.}} If anything trouble
them, they cannot sleep, but fret themselves still, till another object
mitigate, or time wear it out. They have grievous passions, and immoderate
perturbations of the mind, fear, sorrow, \etc{}, yet not so continuate, but
that they are sometimes merry, apt to profuse laughter, which is more to be
wondered at, and that by the authority of \authorfootnote{2631}Galen himself,
by reason of mixture of blood, \li{praerubri jocosis delectantur, et irrisores
plerumque sunt}, if they be ruddy, they are delighted in jests, and oftentimes
scoffers themselves, conceited: and as Rodericus a Vega comments on that place
of Galen, merry, witty, of a pleasant disposition, and yet grievously
melancholy anon after: \li{omnia discunt sine doctore}, saith Aretus, they
learn without a teacher: and as \authorfootnote{2632}Laurentius supposeth,
those feral passions and symptoms of such as think themselves glass, pitchers,
feathers, \etc{}, speak strange languages, \li{a colore cerebri} (if it be in
excess) from the brain's distempered heat.

%SECT. III. MEMB. II. SUBSECT. II.-_Symptoms of windy Hypochondriacal Melancholy_.
\section{Symptoms of windy Hypochondriacal Melancholy.}

"In this hypochondriacal or flatuous melancholy, the symptoms are so
ambiguous," saith \authorfootnote{2633}Crato in a counsel of his for a
noblewoman, "that the most exquisite physicians cannot determine of the part
affected." Matthew Flaccius, consulted about a noble matron, confessed as much,
that in this malady he with Hollerius, Fracastorius, Falopius, and others,
being to give their sentence of a party labouring of hypochondriacal
melancholy, could not find out by the symptoms which part was most especially
affected; some said the womb, some heart, some stomach, \etc{}, and therefore
Crato, \bookcite{\textlatin{consil. 24. lib. 1.}} boldly avers, that in this
diversity of symptoms, which commonly accompany this disease,
\authorfootnote{2634}"no physician can truly say what part is affected." Galen
\bookcite{\textlatin{lib. 3. de loc. affect.}}, reckons up these ordinary
symptoms, which all the Neoterics repeat of Diocles; only this fault he finds
with him, that he puts not fear and sorrow amongst the other signs.
Trincavelius excuseth Diocles, \bookcite{\textlatin{lib. 3. consil. 35.}}
because that oftentimes in a strong head and constitution, a generous spirit,
and a valiant, these symptoms appear not, by reason of his valour and courage.
\authorfootnote{2635}Hercules de Saxonia (to whom I subscribe) is of the same
mind (which I have before touched) that fear and sorrow are not general
symptoms; some fear and are not sad; some be sad and fear not; some neither
fear nor grieve. The rest are these, beside fear and sorrow,
\authorfootnote{2636}"sharp belchings, fulsome crudities, heat in the bowels,
wind and rumbling in the guts, vehement gripings, pain in the belly and stomach
sometimes, after meat that is hard of concoction, much watering of the stomach,
and moist spittle, cold sweat, \li{importunus sudor}, unseasonable sweat all
over the body," as Octavius Horatianus \bookcite{\textlatin{lib. 2. cap. 5.}}
calls it; "cold joints, indigestion, \authorfootnote{2637}they cannot endure
their own fulsome belchings, continual wind about their hypochondries, heat and
griping in their bowels, \li{praecordia sursum convelluntur}, midriff and
bowels are pulled up, the veins about their eyes look red, and swell from
vapours and wind." Their ears sing now and then, vertigo and giddiness come by
fits, turbulent dreams, dryness, leanness, apt they are to sweat upon all
occasions, of all colours and complexions. Many of them are high-coloured
especially after meals, which symptom Cardinal Caecius was much troubled with,
and of which he complained to Prosper Calenus his physician, he could not eat,
or drink a cup of wine, but he was as red in the face as if he had been at a
mayor's feast. That symptom alone vexeth many. \authorfootnote{2638}Some again
are black, pale, ruddy, sometimes their shoulders and shoulder blades ache,
there is a leaping all over their bodies, sudden trembling, a palpitation of
the heart, and that \li{cardiaca passio}, grief in the mouth of the stomach,
which maketh the patient think his heart itself acheth, and sometimes
suffocation, \li{difficultas anhelitus}, short breath, hard wind, strong pulse,
swooning. Montanus \bookcite{\textlatin{consil. 55.}} Trincavelius
\bookcite{\textlatin{lib. 3. consil. 36. et 37.}} Fernelius
\bookcite{\textlatin{cons. 43.}} Frambesarius \bookcite{\textlatin{consult.
lib. 1. consil. 17.}} Hildesheim, Claudinus, \etc{}, give instance of every
particular. The peculiar symptoms which properly belong to each part be these.
If it proceed from the stomach, saith \authorfootnote{2639}Savanarola, 'tis
full of pain wind. Guianerius adds, vertigo, nausea, much spitting, \etc{} If
from the mirach, a swelling and wind in the hypochondries, a loathing, and
appetite to vomit, pulling upward. If from the heart, aching and trembling of
it, much heaviness. If from the liver, there is usually a pain in the right
hypochondry. If from the spleen, hardness and grief in the left hypochondry, a
rumbling, much appetite and small digestion, Avicenna. If from the mesaraic
veins and liver on the other side, little or no appetite, Herc. de Saxonia. If
from the hypochondries, a rumbling inflation, concoction is hindered, often
belching, \etc{} And from these crudities, windy vapours ascend up to the brain
which trouble the imagination, and cause fear, sorrow, dullness, heaviness,
many terrible conceits and chimeras, as Lemnius well observes,
\bookcite{\textlatin{l. 1. c. 16.}} "as \authorfootnote{2640}a black and thick
cloud covers the sun, and intercepts his beams and light, so doth this
melancholy vapour obnubilate the mind, enforce it to many absurd thoughts and
imaginations," and compel good, wise, honest, discreet men (arising to the
brain from the \authorfootnote{2641}lower parts, "as smoke out of a chimney")
to dote, speak, and do that which becomes them not, their persons, callings,
wisdoms. One by reason of those ascending vapours and gripings, rumbling
beneath, will not be persuaded but that he hath a serpent in his guts, a viper,
another frogs. Trallianus relates a story of a woman, that imagined she had
swallowed an eel, or a serpent, and Felix Platerus,
\bookcite{\textlatin{observat. lib. 1.}} hath a most memorable example of a
countryman of his, that by chance, falling into a pit where frogs and frogs'
spawn was, and a little of that water swallowed, began to suspect that he had
likewise swallowed frogs' spawn, and with that conceit and fear, his phantasy
wrought so far, that he verily thought he had young live frogs in his belly,
\li{qui vivebant ex alimento suo}, that lived by his nourishment, and was so
certainly persuaded of it, that for many years afterwards he could not be
rectified in his conceit: He studied physic seven years together to cure
himself, travelled into Italy, France and Germany to confer with the best
physicians about it, and A.D. 1609, asked his counsel amongst the rest; he told
him it was wind, his conceit, \etc{}, but \li{mordicus contradicere, et ore, et
scriptis probare nitebatur}: no saying would serve, it was no wind, but real
frogs: "and do you not hear them croak?" Platerus would have deceived him, by
putting live frog's into his excrements; but he, being a physician himself,
would not be deceived, \li{vir prudens alias, et doctus} a wise and learned man
otherwise, a doctor of physic, and after seven years' dotage in this kind,
\li{a phantasia liberatus est}, he was cured. Laurentius and Goulart have many
such examples, if you be desirous to read them. One commodity above the rest
which are melancholy, these windy flatuous have, \li{lucidia intervalla}, their
symptoms and pains are not usually so continuate as the rest, but come by fits,
fear and sorrow, and the rest: yet in another they exceed all others; and that
is, \authorfootnote{2642}they are luxurious, incontinent, and prone to venery,
by reason of wind, \li{et facile amant, et quamlibet fere amant}. (Jason
Pratensis) \authorfootnote{2643}Rhasis is of opinion, that Venus doth many of
them much good; the other symptoms of the mind be common with the rest.

%SECT. III. MEMB. II. SUBSECT. III.-_Symptoms of Melancholy abounding in the whole body_.
\section[Symptoms In the Whole Body]{Symptoms of Melancholy abounding in the whole body.}

\lettrine{T}{heir} bodies that are affected with this universal melancholy are
most part black, \authorfootnote{2644}"the melancholy juice is redundant all
over," hirsute they are, and lean, they have broad veins, their blood is gross
and thick \authorfootnote{2645}"Their spleen is weak," and a liver apt to
engender the humour; they have kept bad diet, or have had some evacuation
stopped, as haemorrhoids, or months in women, which
\authorfootnote{2646}Trallianus, in the cure, would have carefully to be
inquired, and withal to observe of what complexion the party is of, black or
red. For as Forrestus and Hollerius contend, if \authorfootnote{2647}they be
black, it proceeds from abundance of natural melancholy; if it proceed from
cares, agony, discontents, diet, exercise, \etc{}, they may be as well of any
other colour: red, yellow, pale, as black, and yet their whole blood corrupt:
\li{praerubri colore saepe sunt tales, saepe flavi}, (saith
\authorfootnote{2648}Montaltus \bookcite{\textlatin{cap. 22.}}) The best way to
discern this species, is to let them bleed, if the blood be corrupt, thick and
black, and they withal free from those hypochondriacal symptoms, and not so
grievously troubled with them, or those of the head, it argues they are
melancholy, \li{a toto corpore}. The fumes which arise from this corrupt blood,
disturb the mind, and make them fearful and sorrowful, heavy hearted, as the
rest, dejected, discontented, solitary, silent, weary of their lives, dull and
heavy, or merry, \etc{}, and if far gone, that which Apuleius wished to his
enemy, by way of imprecation, is true in them; \authorfootnote{2649}"Dead men's
bones, hobgoblins, ghosts are ever in their minds, and meet them still in every
turn: all the bugbears of the night, and terrors, fairy-babes of tombs, and
graves are before their eyes, and in their thoughts, as to women and children,
if they be in the dark alone." If they hear, or read, or see any tragical
object, it sticks by them, they are afraid of death, and yet weary of their
lives, in their discontented humours they quarrel with all the world, bitterly
inveigh, tax satirically, and because they cannot otherwise vent their passions
or redress what is amiss, as they mean, they will by violent death at last be
revenged on themselves.

%SECT. III. MEMB. II. SUBSECT. IV.-_Symptoms of Maids, Nuns, and Widows' Melancholy_.
\section{Symptoms of Maids, Nuns, and Widows' Melancholy.}

\lettrine{B}{ecause}  Lodovicus Mercatus in his second book
\bookcite{\textlatin{de mulier. affect. cap. 4.}} and Rodericus a Castro
\bookcite{\textlatin{de morb. mulier. cap. 3. lib. 2.}} two famous physicians
in Spain, Daniel Sennertus of Wittenberg \bookcite{\textlatin{lib. 1. part 2.
cap. 13.}} with others, have vouchsafed in their works not long since
published, to write two just treatises \bookcite{\textlatin{de Melancholia
virginum, Monialium et Viduarum}}, as a particular species of melancholy (which
I have already specified) distinct from the rest; \authorfootnote{2650}(for it
much differs from that which commonly befalls men and other women, as having
one only cause proper to women alone) I may not omit in this general survey of
melancholy symptoms, to set down the particular signs of such parties so
misaffected.

The causes are assigned out of Hippocrates, Cleopatra, Moschion, and those old
\li{Gynaeciorum Scriptores}, of this feral malady, in more ancient maids,
widows, and barren women, \li{ob septum transversum violatum}, saith Mercatus,
by reason of the midriff or \li{Diaphragma}, heart and brain offended with
those vicious vapours which come from menstruous blood, \li{inflammationem
arteriae circa dorsum}, Rodericus adds, an inflammation of the back, which with
the rest is offended by \authorfootnote{2651}that fuliginous exhalation of
corrupt seed, troubling the brain, heart and mind; the brain, I say, not in
essence, but by consent, \li{Universa enim hujus affectus causa ab utero
pendet, et a sanguinis menstrui malitia}, for in a word, the whole malady
proceeds from that inflammation, putridity, black smoky vapours, \etc{}, from
thence comes care, sorrow, and anxiety, obfuscation of spirits, agony,
desperation, and the like, which are intended or remitted; \li{si amatorius
accesserit ardor}, or any other violent object or perturbation of mind. This
melancholy may happen to widows, with much care and sorrow, as frequently it
doth, by reason of a sudden alteration of their accustomed course of life,
\etc{} To such as lie in childbed \li{ob suppressam purgationem}; but to nuns
and more ancient maids, and some barren women for the causes abovesaid, 'tis
more familiar, \li{crebrius his quam reliquis accidit, inquit Rodericus}, the
rest are not altogether excluded.

Out of these causes Rodericus defines it with Areteus, to be \li{angorem
animi}, a vexation of the mind, a sudden sorrow from a small, light, or no
occasion, \authorfootnote{2652}with a kind of still dotage and grief of some
part or other, head, heart, breasts, sides, back, belly, \etc{}, with much
solitariness, weeping, distraction, \etc{}, from which they are sometimes
suddenly delivered, because it comes and goes by fits, and is not so permanent
as other melancholy.

But to leave this brief description, the most ordinary symptoms be these,
\li{pulsatio juxta dorsum}, a beating about the back, which is almost
perpetual, the skin is many times rough, squalid, especially, as Areteus
observes, about the arms, knees, and knuckles. The midriff and heart-strings do
burn and beat very fearfully, and when this vapour or fume is stirred, flieth
upward, the heart itself beats, is sore grieved, and faints, \li{fauces
siccitate praecluduntur, ut difficulter possit ab uteri strangulatione
decerni}, like fits of the mother, \li{Alvus plerisque nil reddit, aliis
exiguum, acre, biliosum, lotium flavum}. They complain many times, saith
Mercatus, of a great pain in their heads, about their hearts, and
hypochondries, and so likewise in their breasts, which are often sore,
sometimes ready to swoon, their faces are inflamed, and red, they are dry,
thirsty, suddenly hot, much troubled with wind, cannot sleep, \etc{} And from
hence proceed \li{ferina deliramenta}, a brutish kind of dotage, troublesome
sleep, terrible dreams in the night, \li{subrusticus pudor et verecundia
ignava}, a foolish kind of bashfulness to some, perverse conceits and opinions,
\authorfootnote{2653}dejection of mind, much discontent, preposterous judgment.
They are apt to loath, dislike, disdain, to be weary of every object, \etc{},
each thing almost is tedious to them, they pine away, void of counsel, apt to
weep, and tremble, timorous, fearful, sad, and out of all hope of better
fortunes. They take delight in nothing for the time, but love to be alone and
solitary, though that do them more harm: and thus they are affected so long as
this vapour lasteth; but by-and-by, as pleasant and merry as ever they were in
their lives, they sing, discourse, and laugh in any good company, upon all
occasions, and so by fits it takes them now and then, except the malady be
inveterate, and then 'tis more frequent, vehement, and continuate. Many of them
cannot tell how to express themselves in words, or how it holds them, what ails
them, you cannot understand them, or well tell what to make of their sayings;
so far gone sometimes, so stupefied and distracted, they think themselves
bewitched, they are in despair, \li{aptae ad fletum, desperationem, dolores
mammis et hypocondriis}. Mercatus therefore adds, now their breasts, now their
hypochondries, belly and sides, then their heart and head aches, now heat, then
wind, now this, now that offends, they are weary of all;
\authorfootnote{2654}and yet will not, cannot again tell how, where or what
offends them, though they be in great pain, agony, and frequently complain,
grieving, sighing, weeping, and discontented still, \li{sine causa manifesta},
most part, yet I say they will complain, grudge, lament, and not be persuaded,
but that they are troubled with an evil spirit, which is frequent in Germany,
saith Rodericus, amongst the common sort: and to such as are most grievously
affected, (for he makes three degrees of this disease in women,) they are in
despair, surely forespoken or bewitched, and in extremity of their dotage,
(weary of their lives,) some of them will attempt to make away themselves. Some
think they see visions, confer with spirits and devils, they shall surely be
damned, are afraid of some treachery, imminent danger, and the like, they will
not speak, make answer to any question, but are almost distracted, mad, or
stupid for the time, and by fits: and thus it holds them, as they are more or
less affected, and as the inner humour is intended or remitted, or by outward
objects and perturbations aggravated, solitariness, idleness, \etc{}

Many other maladies there are incident to young women, out of that one and only
cause above specified, many feral diseases. I will not so much as mention their
names, melancholy alone is the subject of my present discourse, from which I
will not swerve. The several cures of this infirmity, concerning diet, which
must be very sparing, phlebotomy, physic, internal, external remedies, are at
large in great variety in \authorfootnote{2655}Rodericus a Castro, Sennertus,
and Mercatus, which whoso will, as occasion serves, may make use of. But the
best and surest remedy of all, is to see them well placed, and married to good
husbands in due time, \li{hinc illae, lachrymae}, that is the primary cause,
and this the ready cure, to give them content to their desires. I write not
this to patronise any wanton, idle flirt, lascivious or light housewives, which
are too forward many times, unruly, and apt to cast away themselves on him that
comes next, without all care, counsel, circumspection, and judgment. If
religion, good discipline, honest education, wholesome exhortation, fair
promises, fame and loss of good name cannot inhibit and deter such, (which to
chaste and sober maids cannot choose but avail much,) labour and exercise,
strict diet, rigour and threats may more opportunely be used, and are able of
themselves to qualify and divert an ill-disposed temperament. For seldom should
you see an hired servant, a poor handmaid, though ancient, that is kept hard to
her work, and bodily labour, a coarse country wench troubled in this kind, but
noble virgins, nice gentlewomen, such as are solitary and idle, live at ease,
lead a life out of action and employment, that fare well, in great houses and
jovial companies, ill-disposed peradventure of themselves, and not willing to
make any resistance, discontented otherwise, of weak judgment, able bodies, and
subject to passions, (\li{grandiores virgines}, saith Mercatus, \li{steriles et
viduae plerumque melancholicae},) such for the most part are misaffected, and
prone to this disease. I do not so much pity them that may otherwise be eased,
but those alone that out of a strong temperament, innate constitution, are
violently carried away with this torrent of inward humours, and though very
modest of themselves, sober, religious, virtuous, and well given, (as many so
distressed maids are,) yet cannot make resistance, these grievances will
appear, this malady will take place, and now manifestly show itself, and may
not otherwise be helped. But where am I? Into what subject have I rushed? What
have I to do with nuns, maids, virgins, widows? I am a bachelor myself, and
lead a monastic life in a college, \li{nae ego sane ineptus qui haec dixerim},%)
I confess 'tis an indecorum, and as Pallas a virgin blushed, when Jupiter by
chance spake of love matters in her presence, and turned away her face; \li{me
reprimam} though my subject necessarily require it, I will say no more.

And yet I must and will say something more, add a word or two \li{in gratiam
virginum et viduarum}, in favour of all such distressed parties, in
commiseration of their present estate. And as I cannot choose but condole their
mishap that labour of this infirmity, and are destitute of help in this case,
so must I needs inveigh against them that are in fault, more than manifest
causes, and as bitterly tax those tyrannising pseudopoliticians, superstitious
orders, rash vows, hard-hearted parents, guardians, unnatural friends, allies,
(call them how you will,) those careless and stupid overseers, that out of
worldly respects, covetousness, supine negligence, their own private ends
(\li{cum sibi sit interim bene}) can so severely reject, stubbornly neglect,
and impiously contemn, without all remorse and pity, the tears, sighs, groans,
and grievous miseries of such poor souls committed to their charge. How odious
and abominable are those superstitious and rash vows of Popish monasteries, so
to bind and enforce men and women to vow virginity, to lead a single life,
against the laws of nature, opposite to religion, policy, and humanity, so to
starve, to offer violence, to suppress the vigour of youth, by rigorous
statutes, severe laws, vain persuasions, to debar them of that to which by
their innate temperature they are so furiously inclined, urgently carried, and
sometimes precipitated, even irresistibly led, to the prejudice of their soul's
health, and good estate of body and mind: and all for base and private
respects, to maintain their gross superstition, to enrich themselves and their
territories as they falsely suppose, by hindering some marriages, that the
world be not full of beggars, and their parishes pestered with orphans; stupid
politicians; \li{haeccine fieri flagilia}? ought these things so to be carried?
better marry than burn, saith the Apostle, but they are otherwise persuaded.
They will by all means quench their neighbour's house if it be on fire, but
that fire of lust which breaks out into such lamentable flames, they will not
take notice of, their own bowels oftentimes, flesh and blood shall so rage and
burn, and they will not see it: \li{miserum est}, saith Austin, \li{seipsum non
miserescere}, and they are miserable in the meantime that cannot pity
themselves, the common good of all, and \li{per consequens} their own estates.
For let them but consider what fearful maladies, feral diseases, gross
inconveniences, come to both sexes by this enforced temperance, it troubles me
to think of, much more to relate those frequent abortions and murdering of
infants in their nunneries (read \authorfootnote{2656}Kemnitius and others),
and notorious fornications, those Spintrias, Tribadas, Ambubeias, \etc{}, those
rapes, incests, adulteries, \margindef{masturbation}{mastuprations}, sodomies,
buggeries of monks and friars. See Bale's visitation of abbeys,
\authorfootnote{2657}Mercurialis, Rodericus a Castro, Peter Forestus, and
divers physicians; I know their ordinary apologies and excuses for these
things, \li{sed viderint Politici, Medici, Theologi}, I shall more opportunely
meet with them \authorfootnote{2658}elsewhere.


\begin{latin}
\begin{verse}%
\authorfootnote{2659}Illius viduae, aut patronum Virginis hujus,\\*
Ne me forte putes, verbum non amplius addam.\\!
\end{verse}%
\end{latin}



%MEMB. III.

\section{Immediate cause of these precedent Symptoms.}

\lettrine{T}{o} give some satisfaction to melancholy men that are troubled with these
symptoms, a better means in my judgment cannot be taken, than to show them the
causes whence they proceed; not from devils as they suppose, or that they are
bewitched or forsaken of God, hear or see, \etc{} as many of them think, but
from natural and inward causes, that so knowing them, they may better avoid the
effects, or at least endure them with more patience. The most grievous and
common symptoms are fear and sorrow, and that without a cause to the wisest and
discreetest men, in this malady not to be avoided. The reason why they are so,
Aetius discusseth at large, \bookcite{\textlatin{Tetrabib. 2. 2.}} in his first
problem out of Galen, \bookcite{\textlatin{lib. 2. de causis sympt. 1.}} For
Galen imputeth all to the cold that is black, and thinks that the spirits being
darkened, and the substance of the brain cloudy and dark, all the objects
thereof appear terrible, and the \authorfootnote{2660}mind itself, by those
dark, obscure, gross fumes, ascending from black humours, is in continual
darkness, fear, and sorrow; divers terrible monstrous fictions in a thousand
shapes and apparitions occur, with violent passions, by which the brain and
fantasy are troubled and eclipsed. \authorfootnote{2661}Fracastorius,
\bookcite{\textlatin{lib. 2. de intellect}}, "will have cold to be the cause of
fear and sorrow; for such as are cold are ill-disposed to mirth, dull, and
heavy, by nature solitary, silent; and not for any inward darkness (as
physicians think) for many melancholy men dare boldly be, continue, and walk in
the dark, and delight in it:" \li{solum frigidi timidi}: if they be hot, they
are merry; and the more hot, the more furious, and void of fear, as we see in
madmen; but this reason holds not, for then no melancholy, proceeding from
choler adust, should fear. \authorfootnote{2662}Averroes scoffs at Galen for
his reasons, and brings five arguments to repel them: so doth Herc. de Saxonia,
\bookcite{\textlatin{Tract. de Melanch. cap. 3.}} assigning other causes, which
are copiously censured and confuted by Aelianus Montaltus,
\bookcite{\textlatin{cap. 5 and 6.}} Lod. Mercatus \bookcite{\textlatin{de
Inter. morb. cur. lib. 1. cap. 17.}} Altomarus, \bookcite{\textlatin{cap. 7. de
mel.}} Guianerius, \bookcite{\textlatin{tract. 15. c. 1.}} Bright
\bookcite{\textlatin{cap. 37.}} Laurentius, \bookcite{\textlatin{cap. 5.}}
Valesius, \bookcite{\textlatin{med. cont. lib. 5, con. 1.}}
\authorfootnote{2663}"Distemperature," they conclude, "makes black juice,
blackness obscures the spirits, the spirits obscured, cause fear and sorrow."
Laurentius, \bookcite{\textlatin{cap. 13.}} supposeth these black fumes offend
specially the diaphragma or midriff, and so \li{per consequens} the mind, which
is obscured as \authorfootnote{2664}the sun by a cloud. To this opinion of
Galen, almost all the Greeks and Arabians subscribe, the Latins new and old,
\li{internae, tenebrae offuscant animum, ut externae nocent pueris}, as
children are affrighted in the dark, so are melancholy men at all times,
\authorfootnote{2665}as having the inward cause with them, and still carrying
it about. Which black vapours, whether they proceed from the black blood about
the heart, as T. W. Jes. thinks in his treatise of the passions of the mind, or
stomach, spleen, midriff, or all the misaffected parts together, it boots not,
they keep the mind in a perpetual dungeon, and oppress it with continual fears,
anxieties, sorrows, \etc{} It is an ordinary thing for such as are sound to
laugh at this dejected pusillanimity, and those other symptoms of melancholy,
to make themselves merry with them, and to wonder at such, as toys and trifles,
which may be resisted and withstood, if they will themselves: but let him that
so wonders, consider with himself, that if a man should tell him on a sudden,
some of his especial friends were dead, could he choose but grieve? Or set him
upon a steep rock, where he should be in danger to be precipitated, could he be
secure? His heart would tremble for fear, and his head be giddy. P. Byaras,
\bookcite{\textlatin{Tract. de pest.}} gives instance (as I have said)
\authorfootnote{2666}"and put case" (saith he) "in one that walks upon a plank,
if it lie on the ground, he can safely do it: but if the same plank be laid
over some deep water, instead of a bridge, he is vehemently moved, and 'tis
nothing but his imagination, \li{forma cadendi impressa}, to which his other
members and faculties obey." Yea, but you infer, that such men have a just
cause to fear, a true object of fear; so have melancholy men an inward cause, a
perpetual fume and darkness, causing fear, grief, suspicion, which they carry
with them, an object which cannot be removed; but sticks as close, and is as
inseparable as a shadow to a body, and who can expel or overrun his shadow?
Remove heat of the liver, a cold stomach, weak spleen: remove those adust
humours and vapours arising from them, black blood from the heart, all outward
perturbations, take away the cause, and then bid them not grieve nor fear, or
be heavy, dull, lumpish, otherwise counsel can do little good; you may as well
bid him that is sick of an ague not to be a dry; or him that is wounded not to
feel pain.

Suspicion follows fear and sorrow at heels, arising out of the same fountain,
so thinks \authorfootnote{2667}Fracastorius, "that fear is the cause of
suspicion, and still they suspect some treachery, or some secret machination to
be framed against them, still they distrust." Restlessness proceeds from the
same spring, variety of fumes make them like and dislike. Solitariness,
avoiding of light, that they are weary of their lives, hate the world, arise
from the same causes, for their spirits and humours are opposite to light, fear
makes them avoid company, and absent themselves, lest they should be misused,
hissed at, or overshoot themselves, which still they suspect. They are prone to
venery by reason of wind. Angry, waspish, and fretting still, out of abundance
of choler, which causeth fearful dreams and violent perturbations to them, both
sleeping and waking: That they suppose they have no heads, fly, sink, they are
pots, glasses, \etc{} is wind in their heads. \authorfootnote{2668}Herc. de
Saxonia doth ascribe this to the several motions in the animal spirits, "their
dilation, contraction, confusion, alteration, tenebrosity, hot or cold
distemperature," excluding all material humours.
\authorfootnote{2669}Fracastorius "accounts it a thing worthy of inquisition,
why they should entertain such false conceits, as that they have horns, great
noses, that they are birds, beasts," \etc{}, why they should think themselves
kings, lords, cardinals. For the first, \authorfootnote{2670}Fracastorius gives
two reasons: "One is the disposition of the body; the other, the occasion of
the fantasy," as if their eyes be purblind, their ears sing, by reason of some
cold and rheum, \etc{} To the second, Laurentius answers, the imagination
inwardly or outwardly moved, represents to the understanding, not enticements
only, to favour the passion or dislike, but a very intensive pleasure follows
the passion or displeasure, and the will and reason are captivated by
delighting in it.

Why students and lovers are so often melancholy and mad, the philosopher of
\authorfootnote{2671}Conimbra assigns this reason, "because by a vehement and
continual meditation of that wherewith they are affected, they fetch up the
spirits into the brain, and with the heat brought with them, they incend it
beyond measure: and the cells of the inner senses dissolve their temperature,
which being dissolved, they cannot perform their offices as they ought."

Why melancholy men are witty, which Aristotle hath long since maintained in his
problems; and that \authorfootnote{2672}all learned men, famous philosophers,
and lawgivers, \li{ad unum fere omnes melancholici}, have still been
melancholy, is a problem much controverted. Jason Pratensis will have it
understood of natural melancholy, which opinion Melancthon inclines to, in his
book \bookcite{\textlatin{de Anima}}, and Marcilius Ficinus
\bookcite{\textlatin{de san. tuend. lib. 1. cap. 5.}} but not simple, for that
makes men stupid, heavy, dull, being cold and dry, fearful, fools, and
solitary, but mixed with the other humours, phlegm only excepted; and they not
adust, \authorfootnote{2673}but so mixed as that blood he half, with little or
no adustion, that they be neither too hot nor too cold. Aponensis, cited by
Melancthon, thinks it proceeds from melancholy adust, excluding all natural
melancholy as too cold. Laurentius condemns his tenet, because adustion of
humours makes men mad, as lime burns when water is cast on it. It must be mixed
with blood, and somewhat adust, and so that old aphorism of Aristotle may be
verified, \li{Nullum magnum ingenium sine mixtura dementiae}, no excellent wit
without a mixture of madness. Fracastorius shall decide the controversy,
\authorfootnote{2674}"phlegmatic are dull: sanguine lively, pleasant,
acceptable, and merry, but not witty; choleric are too swift in motion, and
furious, impatient of contemplation, deceitful wits: melancholy men have the
most excellent wits, but not all; this humour may be hot or cold, thick, or
thin; if too hot, they are furious and mad: if too cold, dull, stupid,
timorous, and sad: if temperate, excellent, rather inclining to that extreme of
heat, than cold." This sentence of his will agree with that of Heraclitus, a
dry light makes a wise mind, temperate heat and dryness are the chief causes of
a good wit; therefore, saith Aelian, an elephant is the wisest of all brute
beasts, because his brain is driest, \li{et ob atrae, bilis capiam}: this
reason Cardan approves, \bookcite{\textlatin{subtil. l. 12.}} Jo. Baptista
Silvaticus, a physician of Milan, in his first controversy, hath copiously
handled this question: Rulandus in his problems, Caelius Rhodiginus,
\bookcite{\textlatin{lib. 17.}} Valleriola \bookcite{\textlatin{6to. narrat.
med.}} Herc. de Saxonia, \bookcite{\textlatin{Tract. posth. de mel. cap. 3.}}
Lodovicus Mercatus, \bookcite{\textlatin{de inter. morb. cur. lib. cap. 17.}}
Baptista Porta, \bookcite{\textlatin{Physiog. lib. 1. c. 13.}} and many others.

Weeping, sighing, laughing, itching, trembling, sweating, blushing, hearing and
seeing strange noises, visions, wind, crudity, are motions of the body,
depending upon these precedent motions of the mind: neither are tears,
affections, but actions (as Scaliger holds) \authorfootnote{2675}"the voice of
such as are afraid, trembles, because the heart is shaken"
(\bookcite{\textlatin{Conimb. prob. 6. sec. 3. de som.}}) why they stutter or
falter in their speech, Mercurialis and Montaltus, \bookcite{\textlatin{cap.
17.}} give like reasons out of Hippocrates, \authorfootnote{2676}"dryness,
which makes the nerves of the tongue torpid." Fast speaking (which is a symptom
of some few) Aetius will have caused \authorfootnote{2677}"from abundance of
wind, and swiftness of imagination:" \authorfootnote{2678}"baldness comes from
excess of dryness," hirsuteness from a dry temperature. The cause of much
waking in a dry brain, continual meditation, discontent, fears and cares, that
suffer not the mind to be at rest, incontinency is from wind, and a hot liver,
Montanus, \bookcite{\textlatin{cons. 26.}} Rumbling in the guts is caused from
wind, and wind from ill concoction, weakness of natural heat, or a distempered
heat and cold; \authorfootnote{2679}Palpitation of the heart from vapours,
heaviness and aching from the same cause. That the belly is hard, wind is a
cause, and of that leaping in many parts. Redness of the face, and itching, as
if they were flea-bitten, or stung with pismires, from a sharp subtle wind.
\authorfootnote{2680}Cold sweat from vapours arising from the hypochondries,
which pitch upon the skin; leanness for want of good nourishment. Why their
appetite is so great, \authorfootnote{2681}Aetius answers: \li{Os ventris
frigescit}, cold in those inner parts, cold belly, and hot liver, causeth
crudity, and intention proceeds from perturbations, \authorfootnote{2682}our
souls for want of spirits cannot attend exactly to so many intentive
operations, being exhaust, and overswayed by passion, she cannot consider the
reasons which may dissuade her from such affections.

\authorfootnote{2683}Bashfulness and blushing, is a passion proper to men
alone, and is not only caused for \authorfootnote{2684}some shame and ignominy,
or that they are guilty unto themselves of some foul fact committed, but as
\authorfootnote{2685}Fracastorius well determines, \li{ob defectum proprium, et
timorem}, "from fear, and a conceit of our defects; the face labours and is
troubled at his presence that sees our defects, and nature willing to help,
sends thither heat, heat draws the subtlest blood, and so we blush. They that
are bold, arrogant, and careless, seldom or never blush, but such as are
fearful." Anthonius Lodovicus, in his book \bookcite{\textlatin{de pudore}},
will have this subtle blood to arise in the face, not so much for the reverence
of our betters in presence, \authorfootnote{2686}"but for joy and pleasure, or
if anything at unawares shall pass from us, a sudden accident, occurse, or
meeting:" (which Disarius in \authorfootnote{2687}Macrobius confirms) any
object heard or seen, for blind men never blush, as Dandinus observes, the
night and darkness make men impudent. Or that we be staid before our betters,
or in company we like not, or if anything molest and offend us,
\li{erubescentia} turns to \li{rubor}, blushing to a continuate redness.
\authorfootnote{2688}Sometimes the extremity of the ears tingle, and are red,
sometimes the whole face, \li{Etsi nihil vitiosum commiseris}, as Lodovicus
holds: though Aristotle is of opinion, \li{omnis pudor ex vitio commisso}, all
shame for some offence. But we find otherwise, it may as well proceed
\authorfootnote{2689}from fear, from force and inexperience, (so
\authorfootnote{2690}Dandinus holds) as vice; a hot liver, saith Duretus
(\li{notis in Hollerium}:) "from a hot brain, from wind, the lungs heated, or
after drinking of wine, strong drink, perturbations," \etc{}

Laughter what it is, saith \authorfootnote{2691}Tully, "how caused, where, and
so suddenly breaks out, that desirous to stay it, we cannot, how it comes to
possess and stir our face, veins, eyes, countenance, mouth, sides, let
Democritus determine." The cause that it often affects melancholy men so much,
is given by Gomesius, \bookcite{\textlatin{lib. 3. de sale genial. cap. 18.}}
abundance of pleasant vapours, which, in sanguine melancholy especially, break
from the heart, \authorfootnote{2692}"and tickle the midriff, because it is
transverse and full of nerves: by which titillation the sense being moved, and
arteries distended, or pulled, the spirits from thence move and possess the
sides, veins, countenance, eyes." See more in Jossius \bookcite{\textlatin{de
risu et fletu}}, Vives \bookcite{\textlatin{3 de Anima}}. Tears, as Scaliger
defines, proceed from grief and pity, \authorfootnote{2693}"or from the heating
of a moist brain, for a dry cannot weep."

That they see and hear so many phantasms, chimeras, noises, visions, \etc{} as
Fienus hath discoursed at large in his book of imagination, and
\authorfootnote{2694}Lavater \bookcite{\textlatin{de spectris, part. 1. cap. 2.
3. 4.}} their corrupt phantasy makes them see and hear that which indeed is
neither heard nor seen, \li{Qui multum jejunant, aut noctes ducunt insomnes},
they that much fast, or want sleep, as melancholy or sick men commonly do, see
visions, or such as are weak-sighted, very timorous by nature, mad, distracted,
or earnestly seek. \li{Sabini quod volunt somniant}, as the saying is, they
dream of that they desire. Like Sarmiento the Spaniard, who when he was sent to
discover the straits of Magellan, and confine places, by the Prorex of Peru,
standing on the top of a hill, \li{Amaenissimam planitiem despicere sibi visus
fuit, aedificia magnifica, quamplurimos Pagos, alias Turres, splendida Templa},
and brave cities, built like ours in Europe, not, saith mine
\authorfootnote{2695}author, that there was any such thing, but that he was
\li{vanissimus et nimis credulus}, and would fain have had it so. Or as
\authorfootnote{2696}Lod. Mercatus proves, by reason of inward vapours, and
humours from blood, choler, \etc{} diversely mixed, they apprehend and see
outwardly, as they suppose, divers images, which indeed are not. As they that
drink wine think all runs round, when it is in their own brain; so is it with
these men, the fault and cause is inward, as Galen affirms,
\authorfootnote{2697}mad men and such as are near death, \li{quas extra se
videre putant Imagines, intra oculos habent}, 'tis in their brain, which seems
to be before them; the brain as a concave glass reflects solid bodies.
\li{Senes etiam decrepiti cerebrum habent concavum et aridum, ut imaginentur se
videre} (saith \authorfootnote{2698}Boissardus) \li{quae non sunt}, old men are
too frequently mistaken and dote in like case: or as he that looketh through a
piece of red glass, judgeth everything he sees to be red; corrupt vapours
mounting from the body to the head, and distilling again from thence to the
eyes, when they have mingled themselves with the watery crystal which receiveth
the shadows of things to be seen, make all things appear of the same colour,
which remains in the humour that overspreads our sight, as to melancholy men
all is black, to phlegmatic all white, \etc{} Or else as before the organs
corrupt by a corrupt phantasy, as Lemnius, \bookcite{\textlatin{lib. 1. cap.
16.}} well quotes, \authorfootnote{2699}"cause a great agitation of spirits,
and humours, which wander to and fro in all the creeks of the brain, and cause
such apparitions before their eyes." One thinks he reads something written in
the moon, as Pythagoras is said to have done of old, another smells brimstone,
hears Cerberus bark: Orestes now mad supposed he saw the furies tormenting him,
and his mother still ready to run upon him,

\begin{latin}
\begin{verse}%
O mater obsecro noli me persequi\\*
His furiis, aspectu anguineis, horribilibus,\\*
Ecce ecce me invadunt, in me jam ruunt;\\!
\end{verse}%
\end{latin}
\attrib{\getauthornote{2700}}

but Electra told him thus raving in his mad fit, he saw no such sights
at all, it was but his crazed imagination.

\begin{latin}
\begin{verse}%
Quiesce, quiesce miser in linteis tuis,\\*
Non cernis etenim quae videre te putas.\\!
\end{verse}%
\end{latin}
\attrib{\getauthornote{2701}}

So Pentheus (in Bacchis Euripidis) saw two suns, two Thebes, his brain alone
was troubled. Sickness is an ordinary cause of such sights. Cardan,
\bookcite{\textlatin{subtil. 8.}} \li{Mens aegra laboribus et jejuniis fracta,
facit eos videre, audire}, \etc{} And, Osiander beheld strange visions, and
Alexander ab Alexandro both, in their sickness, which he relates
\bookcite{\textlatin{de rerum varietat. lib. 8. cap. 44.}} Albategnius that
noble Arabian, on his death-bed, saw a ship ascending and descending, which
Fracastorius records of his friend Baptista Tirrianus. Weak sight and a vain
persuasion withal, may effect as much, and second causes concurring, as an oar
in water makes a refraction, and seems bigger, bended double, \etc{} The
thickness of the air may cause such effects, or any object not well-discerned
in the dark, fear and phantasy will suspect to be a ghost, a devil, \etc{}
\authorfootnote{2702}\li{Quod nimis miseri timent, hoc facile credunt}, we are
apt to believe, and mistake in such cases. Marcellus Donatus,
\bookcite{\textlatin{lib. 2. cap. 1.}} brings in a story out of Aristotle, of
one Antepharon which likely saw, wheresoever he was, his own image in the air,
as in a glass. Vitellio, \bookcite{\textlatin{lib. 10. perspect.}} hath such
another instance of a familiar acquaintance of his, that after the want of
three or four nights sleep, as he was riding by a river side, saw another
riding with him, and using all such gestures as he did, but when more light
appeared, it vanished. Eremites and anchorites have frequently such absurd
visions, revelations by reason of much fasting, and bad diet, many are deceived
by legerdemain, as Scot hath well showed in his book of the discovery of
witchcraft, and Cardan, \bookcite{\textlatin{subtil. 18.}} suffites, perfumes,
suffumigations, mixed candles, perspective glasses, and such natural causes,
make men look as if they were dead, or with horse-heads, bull's-horns, and such
like brutish shapes, the room full of snakes, adders, dark, light, green, red,
of all colours, as you may perceive in Baptista Porta, Alexis, Albertus, and
others, glow-worms, fire-drakes, meteors, \li{Ignis fatuus}, which Plinius,
\bookcite{\textlatin{lib. 2. cap. 37.}} calls Castor and Pollux, with many such
that appear in moorish grounds, about churchyards, moist valleys, or where
battles have been fought, the causes of which read in Goclenius, Velouris,
Fickius, \etc{} such fears are often done, to frighten children with squibs,
rotten wood, \etc{} to make folks look as if they were dead,
\authorfootnote{2703}\li{solito majores}, bigger, lesser, fairer, fouler,
\li{ut astantes sine capitibus videantur; aut toti igniti, aut forma daemonum,
accipe pilos canis nigri}, \etc{} saith Albertus; and so 'tis ordinary to see
strange uncouth sights by catoptrics: who knows not that if in a dark room, the
light be admitted at one only little hole, and a paper or glass put upon it,
the sun shining, will represent on the opposite wall all such objects as are
illuminated by his rays? with concave and cylinder glasses, we may reflect any
shape of men, devils, antics, (as magicians most part do, to gull a silly
spectator in a dark room), we will ourselves, and that hanging in the air, when
'tis nothing but such an horrible image as \authorfootnote{2704}Agrippa
demonstrates, placed in another room. Roger Bacon of old is said to have
represented his own image walking in the air by this art, though no such thing
appear in his perspectives. But most part it is in the brain that deceives
them, although I may not deny, but that oftentimes the devil deludes them,
takes his opportunity to suggest, and represent vain objects to melancholy men,
and such as are ill affected. To these you may add the knavish impostures of
jugglers, exorcists, mass-priests, and mountebanks, of whom Roger Bacon speaks,
\etc{} \bookcite{\textlatin{de miraculis naturae et artis. cap. 1.}}
\authorfootnote{2705}they can counterfeit the voices of all birds and brute
beasts almost, all tones and tunes of men, and speak within their throats, as
if they spoke afar off, that they make their auditors believe they hear
spirits, and are thence much astonished and affrighted with it. Besides, those
artificial devices to overhear their confessions, like that whispering place of
Gloucester \authorfootnote{2706}with us, or like the duke's place at Mantua in
Italy, where the sound is reverberated by a concave wall; a reason of which
Blancanus in his Echometria gives, and mathematically demonstrates.

So that the hearing is as frequently deluded as the sight, from the same causes
almost, as he that hears bells, will make them sound what he list. "As the fool
thinketh, so the bell clinketh." Theophilus in Galen thought he heard music,
from vapours which made his ears sound, \etc{} Some are deceived by echoes,
some by roaring of waters, or concaves and reverberation of air in the ground,
hollow places and walls. \authorfootnote{2707}At Cadurcum, in Aquitaine, words
and sentences are repeated by a strange echo to the full, or whatsoever you
shall play upon a musical instrument, more distinctly and louder, than they are
spoken at first. Some echoes repeat a thing spoken seven times, as at Olympus,
in Macedonia, as Pliny relates, \bookcite{\textlatin{lib. 36. cap. 15.}} Some
twelve times, as at Charenton, a village near Paris, in France. At Delphos, in
Greece, heretofore was a miraculous echo, and so in many other places. Cardan,
\bookcite{\textlatin{subtil. l. 18}}, hath wonderful stories of such as have
been deluded by these echoes. Blancanus the Jesuit, in his Echometria, hath
variety of examples, and gives his reader full satisfaction of all such sounds
by way of demonstration. \authorfootnote{2708}At Barrey, an isle in the Severn
mouth, they seem to hear a smith's forge; so at Lipari, and those sulphureous
isles, and many such like, which Olaus speaks of in the continent of Scandia,
and those northern countries. Cardan \bookcite{\textlatin{de rerum var. l. 15,
c. 84}}, mentioneth a woman, that still supposed she heard the devil call her,
and speaking to her, she was a painter's wife in Milan: and many such illusions
and voices, which proceed most part from a corrupt imagination.

Whence it comes to pass, that they prophesy, speak several languages, talk of
astronomy, and other unknown sciences to them (of which they have been ever
ignorant): \authorfootnote{2709}I have in brief touched, only this I will here
add, that Arculanus, \bookcite{\textlatin{Bodin. lib. 3, cap. 6, daemon.}} and
some others, \authorfootnote{2710}hold as a manifest token that such persons
are possessed with the devil; so doth \authorfootnote{2711}Hercules de Saxonia,
and Apponensis, and fit only to be cured by a priest. But
\authorfootnote{2712}Guianerius, \authorfootnote{2713}Montaltus, Pomporiatius
of Padua, and Lemnius \bookcite{\textlatin{lib. 2. cap. 2}}, refer it wholly to
the ill-disposition of the \authorfootnote{2714}humour, and that out of the
authority of Aristotle \bookcite{\textlatin{prob. 30. 1}}, because such
symptoms are cured by purging; and as by the striking of a flint fire is
enforced, so by the vehement motion of spirits, they do \li{elicere voces
inauditas}, compel strange speeches to be spoken: another argument he hath from
Plato's \li{reminiscentia}, which all out as likely as that which
\authorfootnote{2715}Marsilius Ficinus speaks of his friend Pierleonus; by a
divine kind of infusion he understood the secrets of nature, and tenets of
Grecian and barbarian philosophers, before ever he heard of, saw, or read their
works: but in this I should rather hold with Avicenna and his associates, that
such symptoms proceed from evil spirits, which take all opportunities of
humours decayed, or otherwise to pervert the soul of man: and besides, the
humour itself is \li{balneum diaboli}, the devil's bath; and as Agrippa proves,
doth entice him to seize upon them.
}
